# Simple Linear Regression

<!-- (C) 2020 Marek Gagolewski, https://www.gagolewski.com -->


## Machine Learning

### What is Machine Learning?

---

An **algorithm** is a well-defined sequence of instructions that,
for a given sequence of input arguments,
yields some desired output.

In other words, it is a specific recipe for a **function**.

Developing algorithms is a tedious task.

In **machine learning**, we build and study computer algorithms
that make *predictions* or *decisions* but which are not
manually programmed.

**Learning** needs some material based upon new knowledge will be acquired.
We need **data**.



### Main Types of Machine Learning Problems

---

Machine Learning Problems include, but are not limited to:

- **Supervised learning** -- for every input point (e.g., a photo)
there is an associated desired output (e.g., whether it depicts a crosswalk)

- **Unsupervised learning** -- inputs are unlabelled, the aim is to discover
the underlying structure in the data (e.g., automatically group customers
w.r.t. common behavioural patterns)

- **Semi-supervised learning** -- some inputs are labelled, the others
are not (definitely a cheaper scenario)


- **Reinforcement learning** -- learn to act based on a post-factum
feedback on the decision made
(e.g., learn to play The Witcher 7)




## Supervised Learning

### Formalism

---

Let $\mathbf{X}=\{\mathfrak{X}^{(1)},\dots,\mathfrak{X}^{(n)}\}$
be an input sample ("a database")
that consists of $n$ objects.



Most often we assume that each object $\mathfrak{X}_i$
is represented using $p$ numbers for some $p$.

We denote this fact as $\mathfrak{X}_i\in \mathbb{R}^p$
(it is *a $p$-dimensional real vector* or
*a sequence of $p$ numbers* or
*a point in a $p$-dimensional real space*
or *an element of a real $p$-space* etc.).

. . .

Of course, this setting is *abstract* in the way that
there might be different realities hidden behind those symbols.

This is what maths is for -- creating *abstractions* or *models*
of complex entities/phenomena so that they can be much more easily manipulated
or understood.

> This is very powerful -- let's spend a moment
contemplating how many real-world situations fit into this framework.

---

If we have "complex" objects on input,
we can try representing them as **feature vectors** (e.g.,
come up with numeric attributes that best describe them in a task at hand).

> How would you represent a patient in a clinic?

> How would you represent a car in an insurance company's database?

> How would you represent a student in an university?

But, e.g., 1920×1080 pixel image
can be "unwound" to a "flat" vector of length 2,073,600.

(\*) There are some algorithms such as Multidimensional Scaling,
Locally Linear Embedding, IsoMap etc.
that can do that automagically.


---




In cases such as this we say that we deal with *structured (tabular) data*\
-- $\mathbf{X}$ can be written as an ($n\times p$)-matrix:
\[
\mathbf{X}=
\left[
\begin{array}{cccc}
x_{1,1} & x_{1,2} & \cdots & x_{1,p} \\
x_{2,1} & x_{2,2} & \cdots & x_{2,p} \\
\vdots & \vdots & \ddots & \vdots \\
x_{n,1} & x_{n,2} & \cdots & x_{n,p} \\
\end{array}
\right]
\]
Mathematically, we denote this as $\mathbf{X}\in\mathbb{R}^{n\times d}$.

> Structured data == think: Excel/Calc spreadsheets, SQL tables etc.




---


**Example**: The famous Fisher's Iris flower dataset, see `?iris` in R
and https://en.wikipedia.org/wiki/Iris_flower_data_set.

```{R}
X <- iris[1:6, 1:4] # first 6 rows and 4 columns
X         # or: print(X)
dim(X)    # gives n and p
dim(iris) # for the full dataset
```

---

$x_{i,j}\in\mathbb{R}$
represents the $j$-th feature of the $i$-th observation,
$j=1,\dots,p$, $i=1,\dots,n$.

For instance:

```{r}
X[3, 2] # 3rd row, 2nd column
```


---

The third observation (data point, row in $\mathbf{X}$)
consists of items $(x_{3,1}, \dots, x_{3,p})$ that can be extracted by calling:

```{R}
X[3,]
as.numeric(X[3,]) # drops names
```

```{R}
length(X[3,])
```

---

Moreover, the second feature/variable/column
is comprised of
$(x_{1,2}, x_{2,2}, \dots, x_{n,2})$:

```{r}
X[,2]
length(X[,2])
```


---

We will sometimes use the following notation to emphasise that
the $\mathbf{X}$ matrix  consists of $n$ rows
or $p$ columns:

\[
\mathbf{X}=\left[
\begin{array}{c}
\mathbf{x}_{1,\cdot} \\
\mathbf{x}_{2,\cdot} \\
\vdots\\
\mathbf{x}_{n,\cdot} \\
\end{array}
\right]
=
\left[
\begin{array}{cccc}
\mathbf{x}_{\cdot,1} &
\mathbf{x}_{\cdot,2} &
\cdots &
\mathbf{x}_{\cdot,p} \\
\end{array}
\right].
\]

---

Here, $\mathbf{x}_{i,\cdot}$ is a *row vector* of length $p$,
i.e., a $(1\times p)$-matrix:

\[
\mathbf{x}_{i,\cdot} = \left[
\begin{array}{cccc}
x_{i,1} &
x_{i,2} &
\cdots &
x_{i,p} \\
\end{array}
\right].
\]

Moreover, $\mathbf{x}_{\cdot,j}$ is a *column vector* of length $n$,
i.e., an $(n\times 1)$-matrix:

\[
\mathbf{x}_{\cdot,j} = \left[
\begin{array}{cccc}
x_{1,j} &
x_{2,j} &
\cdots &
x_{n,j} \\
\end{array}
\right]^T=\left[
\begin{array}{c}
{x}_{1,j} \\
{x}_{2,j} \\
\vdots\\
{x}_{n,j} \\
\end{array}
\right],
\]

where $\cdot^T$ denotes the *transpose* of a given matrix --
think of this as a kind of rotation; it allows us to introduce a set of
"vertically stacked" objects using a single inline formula.



### Desired Outputs

---

In supervised learning,
apart from the inputs we are also given the corresponding
reference/desired outputs.

> The aim of supervised learning is to try to create an "algorithm" that,
given an input point, generates an output that is as close to the desired one
as possible. The given data sample will be used to "train" this "model".

Usually the reference outputs are encoded as single numbers (scalars)
or textual labels.

In other words, with each input $\mathbf{x}_{i,\cdot}$ we associate
the desired output $y_i$:

```{r, echo=-1}
set.seed(123)
# in iris, iris[, 5] gives the Ys
iris[sample(nrow(iris), 3), ]  # three random rows
```

---


Hence, our dataset is $[\mathbf{X}\ \mathbf{y}]$ --
where each object is represented as a row vector
$[\mathbf{x}_{i,\cdot}\ y_i]$, $i=1,\dots,n$:

\[
[\mathbf{X}\ \mathbf{y}]=
\left[
\begin{array}{ccccc}
x_{1,1} & x_{1,2} & \cdots & x_{1,p} & y_1\\
x_{2,1} & x_{2,2} & \cdots & x_{2,p} & y_2\\
\vdots & \vdots & \ddots & \vdots    & \vdots\\
x_{n,1} & x_{n,2} & \cdots & x_{n,p} & y_n\\
\end{array}
\right],
\]

where

\[
\mathbf{y} = \left[
\begin{array}{cccc}
y_{1} &
y_{2} &
\cdots &
y_{n} \\
\end{array}
\right]^T=\left[
\begin{array}{c}
{y}_{1} \\
{y}_{2} \\
\vdots\\
{y}_{n} \\
\end{array}
\right].
\]


### Types of Supervised Learning Problems

---

Depending on the type of the elements in $\mathbf{y}$
(the domain of $\mathbf{y}$),
supervised learning problems are usually
classified as:

- **regression** -- each $y_i$ is a real number

    > e.g., $y_i=$ future market stock price
    with $\mathbf{x}_{i,\cdot}=$ prices from $p$ previous days

- **classification** -- each $y_i$ is a discrete label

    > e.g., $y_i=$ 0/healthy or 1/ill
    with $\mathbf{x}_{i,\cdot}=$ a patient's health data


- **ordinal regression** (a.k.a. ordinal classification) -- each $y_i$ is a rank

    > e.g., $y_i=$ rating of a product on the scale 1--5
    with $\mathbf{x}_{i,\cdot}=$ ratings of $p$ most similar products


---

Example Problems -- Discussion:

> Which of the following are instances of classification problems? Which of them are regression tasks?
>
> What kind of data should you gather in order to tackle them?
>
> - Email spam detection
> - Market stock price prediction
> - Predict the likeability of a new ad
> - Credit risk assessment
> - Detect existence of tumour tissues in medical images
> - Predict time-to-recovery of cancer patients
> - Recognise smiling faces on photographs
> - Detect unattended luggage in airport security camera footage
> - Turn on emergency braking to avoid collisions with pedestrians

---

A single dataset can become an instance of many different ML problems.

Examples -- the `wines` dataset:

```{r,echo=-1}
set.seed(123)
wines <- read.csv("datasets/winequality-all.csv", comment="#")
wines[1,]
```

---


```{r,echo=-1}
par(mar=c(4,4,1,1))
summary(wines$alcohol) # continuous variable
hist(wines$alcohol, las=1, main=""); box()
```

---


```{r,echo=-1}
par(mar=c(4,4,1,1))
table(wines$color) # binary variable
barplot(table(wines$color), las=1)
```

---


```{r,echo=-1}
par(mar=c(4,4,1,1))
table(wines$response) # ordinal variable
barplot(table(wines$response), las=1)
```




## Simple Regression

### Introduction

---

**Simple regression** is the easiest setting to start with -- let's assume
$p=1$, i.e., all inputs are 1-dimensional.
Denote $x_i=x_{i,1}$.

We will use it to build many intuitions, for example, it'll be easy
to illustrate the concepts graphically.

```{r,echo=-1}
par(mar=c(4,4,1,1))
library("ISLR") # Credit dataset
plot(Credit$Balance, Credit$Rating, las=1) # scatter plot
```

---

In what follows we will be modelling the Credit Rating ($Y$)
as a function of the average Credit Card Balance ($X$) in USD
for customers with positive Balance only.

```{r}
X <- as.matrix(as.numeric(Credit$Balance[Credit$Balance>0]))
Y <- as.matrix(as.numeric(Credit$Rating[Credit$Balance>0]))
```

---

```{r,echo=-1}
par(mar=c(4,4,1,1))
plot(X, Y, las=1)
```

Our aim is to construct a function $f$ that
**models** Rating as a function of Balance,
$f(X)=Y$.

We are equipped with $n=`r length(X)`$ reference (observed) Ratings
$\mathbf{y}=[y_1\ \cdots\ y_n]^T$
for particular Balances $\mathbf{x}=[x_1\ \cdots\ x_n]^T$.

---


Note the following naming conventions:

* Variable types:

    - $X$ -- independent/explanatory/predictor variable

    - $Y$ -- dependent/response/predicted variable

* Also note that:

    - $Y$ -- idealisation (any possible Rating)

    - $\mathbf{y}=[y_1\ \cdots\ y_n]^T$ -- values actually observed




---

The model will not be ideal, but it might be usable:

- We will be able to **predict** the rating of any new client.


    > What should be the rating of a client with Balance of 1500?\
    > What should be the rating of a client with Balance of 2500?

- We will be able to **describe** (understand) this reality using a single mathematical formula
so as to infer that there is an association between $X$ and $Y$

    > Think of "data compression" and laws of physics, e.g., $E=mc^2$
    or ${\displaystyle i\hbar {\frac {d}{dt}}\vert \Psi (t)\rangle ={\hat {H}}\vert \Psi (t)\rangle }$




---


Mathematically, we will assume that there is some "true" function that models data
(true relationship between $Y$ and $X$),
but the observed outputs are subject to **additive error**:
\[Y=f(X)+\varepsilon.\]


$\varepsilon$ is a random term, classically we assume that
errors are independent of each other,
have expected value of $0$ (there is no systematic error = unbiased)
and that they follow a normal distribution.

---

We denote this as $\varepsilon\sim\mathcal{N}(0, \sigma)$
(read: random variable $\varepsilon$ follows a normal distribution
with expect value of $0$ and standard deviation of $\sigma$ for some $\sigma\ge 0$).

$\sigma$ controls the amount of noise (and hence, uncertainty).
Here is the plot of the probability distribution function (PDFs, densities)
of $\mathcal{N}(0, \sigma)$ for different $\sigma$s:

```{r,echo=FALSE, echo=FALSE, message=FALSE}
par(mar=c(3,3,1,1))
y <- seq(-3,3,length.out=101)
plot(y, dnorm(y, mean=0), type='l', ann=FALSE, las=1, ylim=c(0,dnorm(0,0,0.5)))
lines(y, dnorm(y, mean=0, sd=0.5), type='l', col=2, lty=2)
legend("topleft", lty=c(1,2), col=c(1,2), legend=expression(sigma*"="*1, sigma*"="*0.5))
abline(v=0)
```


### Search Space and Objective

---

There are many different functions that can be **fitted** into
the observed $(\mathbf{x},\mathbf{y})$

```{r, echo=FALSE, message=FALSE}
par(mar=c(4,4,1,1))
plot(X, Y, las=1, col="#00000044")

#library("mda")
#f <- mars(as.matrix(X), Y, degree=5)
#x <- seq(min(X), max(X), length.out=101)
#y <- predict(f, x)
#lines(x,y,col=4,lty=4,lwd=2)

f <- lm(Y~poly(X, 2))
x <- seq(min(X), max(X), length.out=101)
y <- predict(f, data.frame(X=x))
lines(x,y,col=4,lty=1,lwd=3)

f <- lm(Y~poly(X, 15))
x <- seq(min(X), max(X), length.out=101)
y <- predict(f, data.frame(X=x))
lines(x,y,col=3,lty=1,lwd=3)

lines(X[order(X)], Y[order(Y)],col=6,lty=1,lwd=3)

abline(lm(Y~X), col=2,lwd=3)

```

Thus, we need a **model selection criterion**.



---


Usually, we will be interested in a model
that minimises appropriately aggregated **residuals**
$f(x_i)-y_i$, i.e.,
**predicted outputs minus observed outputs**,
often denoted with $\hat{y}_i-y_i$.

```{r, echo=FALSE, message=FALSE}
par(mar=c(4,4,1,1))
plot(X, Y, las=1, col=1, xlim=c(1500,2000), ylim=c(400, 1000))
f <- lm(Y~poly(X, 2))
Y2 <- predict(f, data.frame(X=X))
points(X, Y2, pch=4, col=2)
x <- seq(min(X), max(X), length.out=101)
y <- predict(f, data.frame(X=x))
lines(x,y,col=2,lty=1,lwd=2)
segments(X, Y, X, Y2, lty=2)
legend("bottomright", legend=c("observed Y", "predicted Y"), pch=c(1,4), col=c(1, 2))
```

---


Top choice: sum of squared residuals:
\[
\begin{array}{rl}
\mathrm{SSR}(f|\mathbf{x},\mathbf{y})
& = \left( f(x_1)-y_1 \right)^2 + \dots + \left( f(x_n)-y_n \right)^2 \\
& =
\displaystyle\sum_{i=1}^n \left( f(x_i)-y_i \right)^2
\end{array}
\]

> Read "$\sum_{i=1}^n z_i$" as "the sum of $z_i$ for $i$ from $1$ to $n$";
this is just a shorthand for $z_1+z_2+\dots+z_n$.

> The notation $\mathrm{SSR}(f|\mathbf{x},\mathbf{y})$ means that
it is the error measure
corresponding to the model $(f)$ *given* our data.\
> We could've  denoted it  with $\mathrm{SSR}_{\mathbf{x},\mathbf{y}}(f)$
or even $\mathrm{SSR}(f)$ to emphasise that $\mathbf{x},\mathbf{y}$
are just fixed  values and we are not  interested
in changing them at all.

---

We enjoy SSR because (amongst others):

- larger errors are penalised much more than smaller ones

    > (this can be considered a drawback as well)



- (\*\*) statistically speaking, this has a clear underlying interpretation

    > (assuming errors are normally distributed,
    finding a model minimising the SSR is equivalent
    to maximum likelihood estimation)

- the models minimising the SSR can often be found easily

    > (corresponding optimisation tasks have an analytic solution --
    studied already by Gauss in the late 18th century)


(\*\*) Other choices:

- regularised SSR, e.g., lasso or ridge regression (in the case of multiple input variables)
- sum or median of absolute values (robust regression)



---



Fitting a model to data can be written as an optimisation problem:

\[
\min_{f\in\mathcal{F}} \mathrm{SSR}(f|\mathbf{x},\mathbf{y}),
\]

i.e., find $f$ minimising the SSR **(seek "best" $f$)**\
amongst the set of admissible models $\mathcal{F}$.


Example $\mathcal{F}$s:

- $\mathcal{F}=\{\text{All possible functions of one variable}\}$  --  if there are no repeated
$x_i$'s, this corresponds to data *interpolation*; note that there
are many functions that give SSR of $0$.

- $\mathcal{F}=\{ x\mapsto x^2, x\mapsto \cos(x), x\mapsto \exp(2x+7)-9 \}$ -- obviously
an ad-hoc choice but you can easily choose the best amongst the 3 by computing 3 sums of squares.

- $\mathcal{F}=\{ x\mapsto ax+b\}$ -- the space of linear functions of one variable

- etc.


(e.g., $x\mapsto x^2$ is read "$x$ maps to $x^2$" and is
an elegant way to define an inline function $f$ such that $f(x)=x^2$)




### Simple Linear Regression

---

If the family of admissible models $\mathcal{F}$ consists only of all linear functions of one variable,
we deal with a **simple linear regression**.

Our problem becomes:


\[
\min_{a,b\in\mathbb{R}} \sum_{i=1}^n \left(
ax_i+b-y_i
\right)^2
\]


In other words, we seek best fitting line in terms of the squared residuals.

This is the **method of least squares**.

This is particularly nice, because our search space
is just $\mathbb{R}^2$ -- easy to handle both analytically and numerically.

---

```{r,echo=FALSE,fig.height=6}
par(mar=c(4,4,1,1))
plot(X, Y, las=1, col="#00000044")
m1 <- lm(Y~X)
abline(m1, col=2,lwd=3)
m2 <- lm(Y~X+0)
abline(m2, col=3,lwd=3)
m3 <- L1pack::lad(Y~X)
abline(m3, col=4,lwd=3)
legend("topleft",
    lty=1,
    col=c(2,3,4),
    legend=c(
        sprintf("y=%.3fx+%03.0f; SSR=%03g",m1$coefficients[2],m1$coefficients[1], sum(m1$residuals^2)),
        sprintf("y=%.3fx+%03.0f; SSR=%03g",m2$coefficients[1],0, sum(m2$residuals^2)),
        sprintf("y=%.3fx+%03.0f; SSR=%03g",m3$coefficients[2],m3$coefficients[1], sum(m3$residuals^2))
    ))
```

> Which of these is the least squares solution?


### Solution in R

---

Let's fit the linear model minimising the SSR in R.
For convenience, let us store both $\mathbf{x}$ and $\mathbf{y}$
in a data frame:

```{r}
XY <- data.frame(X=X, Y=Y)
head(XY, 3)
```

The `lm()` function (`l`inear `m`odels) has a convenient *formula*-based interface.

> In R, the expression "`Y~X`" denotes a formula, which we read as:
variable `Y` is a function of `X`.
Note that the dependent variable is on the left side of the formula.



```{r}
f <- lm(Y~X, data=XY)
```

> Note that here `X` and `Y` refer to column names in the `XY` data frame.

---

```{r}
print(f)
```

Hence, the fitted model is:
\[
Y = f(X) = `r f$coefficient[2]`X+`r f$coefficient[1]`
\qquad (+ \varepsilon)
\]

---

Coefficient $a$ (slope):

```{r}
f$coefficient[2]
```

Coefficient $b$ (intercept):

```{r}
f$coefficient[1]
```

---

SSR:

```{r}
sum(f$residuals^2)
sum((f$coefficient[2]*X+f$coefficient[1]-Y)^2) # equivalent
```

---

To make a prediction:

```{r}
Xnew <- data.frame(X=c(1500, 2000, 2500))
f$coefficient[2]*Xnew$X + f$coefficient[1]
```


or:

```{r}
predict(f, Xnew)
```

---

However:

```{r}
predict(f, data.frame(X=c(5000)))
```

This is more than the highest possible rating -- we have been extrapolating
way beyond the observable data range.

---

Plotting:

```{r,echo=-1}
par(mar=c(4,4,1,1))
plot(X, Y, col="#000000aa", las=1)
abline(f, col=2, lwd=3)
```




---


Note that our $Y=aX+b$ model is **interpretable**
and **well-behaving**:

> (not all machine learning models will have this feature, think: deep neural networks,
which we rather conceive as *black boxes*)

* we know that by increasing $X$ by a small amount,
$Y$ will also increase (positive correlation),

* the model is continuous -- small change in $X$
doesn't yield any drastic change in $Y$,

* we know what will happen if we increase or decrease $X$ by, say, $100$,

* the function is invertible -- if we want Rating of $500$,
we can compute the associated preferred Balance that should yield it
(provided that the model is valid).




### Derivation of the Solution (\*\*)

---

> (You can safely skip this part if you are yet to know
how to search for a minimum of a function of many variables
and what are partial derivatives)

Denote with:

\[
E(a,b)=\mathrm{SSR}(x\mapsto ax+b|\mathbf{x},\mathbf{y})
=\sum_{i=1}^n \left( ax_i+b - y_i \right) ^2.
\]

We seek the minimum of $E$ w.r.t. both $a,b$.



Theorem.

: If $E$ has a (local) minimum at $(a^*,b^*)$,
then its partial derivatives vanish therein,
i.e., $E_a'(a^*, b^*) = 0$ and $E_b'(a^*, b^*)=0$.


---

We have:

\[
E(a,b)   = \displaystyle\sum_{i=1}^n \left( ax_i+b - y_i \right) ^2.
\]

We need to compute the partial derivatives $\partial E/\partial a$ (derivative of $E$
w.r.t. variable $a$ -- all other terms treated as constants)
and $\partial E/\partial b$ (w.r.t. $b$).

Useful rules -- derivatives w.r.t. $a$ (denote $f'(a)=(f(a))'$):

- $(f(a)+g(a))'=f'(a)+g'(a)$ (derivative of sum is sum of derivatives)
- $(f(a) g(a))' = f'(a)g(a) + f(a)g'(a)$ (derivative of product)
- $(f(g(a)))' = f'(g(a)) g'(a)$ (chain rule)
- $(c)' = 0$ for any constant $c$ (expression not involving $a$)
- $(a^p)' = pa^{p-1}$ for any $p$
- in particular: $(c a^2+d)'=2ca$, $(ca)'=c$, $((ca+d)^2)'=2(ca+d)c$ (application of the above rules)


<!--\[
\begin{array}{rl}
E_a'(a,b)=&2\displaystyle\sum_{i=1}^n \left( ax_i+b - y_i \right) x_i \\
E_b'(a,b)=&2\displaystyle\sum_{i=1}^n \left( ax_i+b - y_i \right) \\
\end{array}
\]-->

We seek $a,b$ such that $\frac{\partial E}{\partial a}(a,b) = 0$
and $\frac{\partial E}{\partial b}(a,b)=0$.

---

\[
\left\{
\begin{array}{rl}
\frac{\partial E}{\partial a}(a,b)=&2\displaystyle\sum_{i=1}^n \left( ax_i+b - y_i \right) x_i = 0 \\
\frac{\partial E}{\partial b}(a,b)=&2\displaystyle\sum_{i=1}^n \left( ax_i+b - y_i \right) = 0\\
\end{array}
\right.
\]

This is a system of 2 linear equations. Easy.

Rearranging like back in the school days:

\[
\left\{
\begin{array}{rl}
a \displaystyle\sum_{i=1}^n x_i x_i + b \displaystyle\sum_{i=1}^n x_i = & \displaystyle\sum_{i=1}^n x_i y_i \\
a \displaystyle\sum_{i=1}^n x_i+ b n = & \displaystyle\sum_{i=1}^n  y_i  \\
\end{array}
\right.
\]

---

It is left as an exercise to show that the solution is:

\[
\left\{
\begin{array}{rl}
a^*  = & \dfrac{
n \displaystyle\sum_{i=1}^n x_i y_i - \displaystyle\sum_{i=1}^n  y_i \displaystyle\sum_{i=1}^n x_i
}{
n \displaystyle\sum_{i=1}^n x_i x_i -   \displaystyle\sum_{i=1}^n x_i\displaystyle\sum_{i=1}^n x_i
}\\
b^* = & \dfrac{1}{n}\displaystyle\sum_{i=1}^n  y_i - a^*  \dfrac{1}{n} \displaystyle\sum_{i=1}^n x_i  \\
\end{array}
\right.
\]



(we should additionally perform the second derivative test
to assure that this is the minimum of $E$ -- which is exactly the case though)

<!-- TODO rewrite nicely with means and correlation coefficients etc.

Pearson's r introduced in the next chapter though

(a <- cor(x,y)*sd(y)/sd(x))
(b <- mean(y)-a*mean(x))
-->



---

Sanity check:

```{r}
n <- length(X)
a <- (n*sum(X*Y)-sum(X)*sum(Y))/(n*sum(X*X)-sum(X)^2)
b <- mean(Y)-a*mean(X)
c(a, b) # the same as f$coefficients
```


---

(\*\*) In the next chapter, we will introduce the notion of Pearson's
linear coefficient, $r$ (see `cor()` in R). It might be shown that
$a$ and $b$ can also be rewritten as:

```{r}
(a <- cor(X,Y)*sd(Y)/sd(X))
(b <- mean(Y)-a*mean(X))
```

<!-- TODO show optim() -->






##  Vector and Matrix Algebra Basics

### Motivation

---

Vector and matrix algebra provides us with a convenient language for
expressing computations on tabular data.

Vector and matrix algebra operations are supported by every
major programming language -- either natively (e.g., R, Matlab, GNU Octave, Mathematica)
or via an additional library/package (e.g, Python with numpy, tensorflow or pytorch;
C++ with Eigen/Armadillo; C, C++ or Fortran with LAPACK).

Using matrix notation leads to a more concise and readable code.
It might also be faster to compute.

---

For instance, given two vectors $\boldsymbol{x}=(x_1,\dots,x_n)$
and $\boldsymbol{y}=(y_1,\dots,y_n)$ like:

```{r}
x <- c(1.5, 3.5, 2.3,-6.5)
y <- c(2.9, 8.2,-0.1, 0.8)
```

Instead of writing:

```{r}
s <- 0
n <- length(x)
for (i in 1:n)
    s <- s + (x[i]-y[i])^2
sqrt(s/n)
```

to mean:

\[
\sqrt{\frac{1}{n} \sum_{i=1}^n (x_i-y_i)^2}
\]

---

we'd rather write:

```{r}
sqrt(mean((x-y)^2))
```

or even:

\[
\frac{1}{\sqrt{n}} \|\boldsymbol{x}-\boldsymbol{y}\|_2
\]


In order to be able to read such a notation,
we only have to get to know the "building blocks". There are just a few
of them, but it takes some time to become comfortable with them.


---

Also note that vectorised code is much faster and much more readable than the `for` loop-based one:

```{r,cache=TRUE,benchmarkvectorised}
library("microbenchmark")
x <- runif(10000) # 10000 random numbers in [0,1]
y <- runif(10000)
print(microbenchmark(
    t1={s <- 0; n <- length(x);
        for (i in 1:n) s <- s + (x[i]-y[i])^2; sqrt(s/n)},
    t2=sqrt(mean((x-y)^2))
), signif=3, unit='relative')
```





### Vector-Scalar Operations

---

Vector-scalar arithmetic operations
such as $s\boldsymbol{x}$ (multiplication of a vector $\boldsymbol{x}=(x_1,\dots, x_n)$
by a scalar $s$) result in a vector $\boldsymbol{y}$ such that $y_i=s x_i$, $i=1,\dots,n$.

The same rule holds for, e.g., $s+\boldsymbol{x}$, $\boldsymbol{x}-s$, $\boldsymbol{x}/s$.


```{r}
0.5 * c(1, 10, 100)
10 + 1:5
seq(0, 10, by=2)/10
```

---

By $-\boldsymbol{x}$ we will mean $(-1)\boldsymbol{x}$:

```{r}
-seq(0, 1, length.out=5)
```

---

Note that in R the same rule applies for exponentiation:

```{r}
(0:5)^2 # synonym: (1:5)**2
2^(0:5)
```

However, in mathematics, we are **not** used  to writing
$2^{\boldsymbol{x}}$ or $\boldsymbol{x}^2$.



### Vector-Vector Operations

---

Let $\boldsymbol{x}=(x_1,\dots,x_n)$ and $\boldsymbol{y}=(y_1,\dots,y_n)$ be two vectors of identical lengths.

Arithmetic operations $\boldsymbol{x}+\boldsymbol{y}$ and $\boldsymbol{x}-\boldsymbol{y}$ are performed *elementwise*,
i.e., they result in a vector $\boldsymbol{z}$ such that
$z_i=x_i+y_i$ and $z_i=x_i-y_i$, respectively, $i=1,\dots,n$.

```{r}
x <- c(1,  2,   3,    4)
y <- c(1, 10, 100, 1000)
x+y
x-y
```

---

Although in mathematics we are **not** used to using any special notation
for elementwise multiplication, division and exponentiation, this is available in R.

```{r}
x*y
x/y
y^x
```

---

Moreover, in R the **recycling rule** is applied if we perform elementwise
operations on vectors of *different* lengths -- the shorter
vector is recycled as many times as needed to match the length of the longer
vectors, just as if we were performing:

```{r}
rep(1:3, length.out=12) # recycle 1,2,3 to get 12 values
```

---

```{r}
1:6 * c(1)
1:6 * c(1,10)
1:6 * c(1,10,100)
1:6 * c(1,10,100,1000)
```

Note that a warning is not an error  -- we still get a sensible result.


---

In R:

* comparison operators such as `<` (less than), `<=` (less than or equal),
`==` (equal), `!=` (not equal), `>` (greater than) and `>=` (greater than or equal)
as well as
* logical operators like `&` (and) and `|` (or)

are performed in the same manner as above, i.e.:

* they are elementwise operations and
* recycling rule is applied if necessary.

However, they return *logical* vectors in result.

---


```{r}
1:2 == 1:4 # c(1,2,1,2) == c(1,2,3,4)
z <- c(0, 3, -1, 1, 0.5)
(z >= 0) & (z <= 1)
```

Also note that in R there are 3 (!) logical values:
`TRUE`, `FALSE` and `NA` (not available, missing, null).


---



Generally, operations on `NA`s yield `NA` in result unless other solution
makes sense.

```{r}
c(0, NA, 2)*c(1, 10, 100)
u <- c(TRUE, FALSE, NA)
v <- c(TRUE, TRUE, TRUE, FALSE, FALSE, FALSE, NA, NA, NA)
u & v # elementwise AND (conjunction)
u | v # elementwise OR  (disjunction)
!u    # elementwise NOT (negation)
```

### Other Vector Operations

---

R implement a couple of *aggregation* functions:

* `sum(x)` = $\sum_{i=1}^n x_i=x_1+x_2+\dots+x_n$
* `prod(x)` = $\prod_{i=1}^n x_i=x_1 x_2 \dots x_n$
* `mean(x)` = $\frac{1}{n}\sum_{i=1}^n x_i$ -- arithmetic mean
* `var(x)` = `sum((x-mean(x))^2)/(length(x)-1)` =
$\frac{1}{n-1} \sum_{i=1}^n \left(x_i - \frac{1}{n}\sum_{j=1}^n x_j \right)^2$
-- variance
* `sd(x)` = `sqrt(var(x))` -- standard deviation

see also: `min()`, `max()`, `median()`, `quantile()`.

> Remember that you can always access the R manual by typing
`?functionname`, e.g., `?quantile`.

---

Mathematically, we will also be interested in the following norms:

* Euclidean norm:
\[
\|\boldsymbol{x}\| = \|\boldsymbol{x}\|_2 = \sqrt{ \sum_{i=1}^n x_i^2 }
\]
this is nothing else than the *length* of the vector $\boldsymbol{x}$
* Manhattan (taxicab) norm:
\[
\|\boldsymbol{x}\|_1 = \sum_{i=1}^n |x_i|
\]
* Chebyshev (maximum) norm:
\[
\|\boldsymbol{x}\|_\infty = \max_{i=1,\dots,n} |x_i|
= \max\{ |x_1|, |x_2|, \dots, |x_n| \}
\]

---

```{r}
z <- c(1, 2)
sqrt(sum(z^2)) # or norm(z, "2"); Euclidean
sum(abs(z))    # Manhattan
max(abs(z))    # Chebyshev
```

---

Also note that:

\[
\| \boldsymbol{x}-\boldsymbol{y} \| = \sqrt{
\sum_{i=1}^n \left(x_i-y_i\right)^2
}
\]

gives the *Euclidean distance* (metric) between the two vectors.

```{r}
u <- c(1, 0)
v <- c(1, 1)
sqrt(sum((u-v)^2))
```


---

What is more, given two vectors of identical lengths,
$\boldsymbol{x}$ and $\boldsymbol{y}$,
we define their *dot product* (a.k.a. *scalar, inner product*) as:

\[
\boldsymbol{x}\cdot\boldsymbol{y} = \sum_{i=1}^n x_i y_i.
\]

> This is not the same as elementwise vector multiplication in R.

```{r}
u <- c(1, 0)
v <- c(1, 1)
sum(u*v)
```

> (\*) Note that the squared Euclidean norm of a vector is equal to the dot
product of the vector and itself,
$\|\boldsymbol{x}\|^2 = \boldsymbol{x}\cdot\boldsymbol{x}$.

---

Interestingly, a dot product has a nice geometrical interpretation:

\[
\boldsymbol{x}\cdot\boldsymbol{y} = \|\boldsymbol{x}\| \|\boldsymbol{y}\|
\cos\alpha
\]
where $\alpha$ is the angle between the two vectors.

> Read: it is the product of the lengths of the two vectors
and the cosine of the angle between them.

> You can get the cosine part by computing the dot product of the *normalised*
vectors, i.e., such that their lengths are equal to 1.

---


```{r,echo=FALSE}
par(mar=c(2.5,2,0.5,0.5))
plot(rbind(u,v), xlim=c(0,1), ylim=c(0,1), las=1, ann=FALSE, asp=1)
arrows(0, 0, u[1], u[2])
arrows(0, 0, v[1], v[2])
text(u[1], u[2], sprintf("[%d,%d]", u[1], u[2]), pos=4)
text(v[1], v[2], sprintf("[%d,%d]", v[1], v[2]), pos=4)
```

```{r}
len_u <- sqrt(sum(u^2)); len_v <- sqrt(sum(v^2))
(cos_angle_uv <- (sum(u*v)/(len_u*len_v)))
acos(cos_angle_uv)*180/pi # angle in degs
```


---

Furthermore, R supports numerous mathematical functions, e.g.,
`sqrt()`, `abs()`, `round()`, `log()`, `exp()`, `cos()`, `sin()`.

All of them are vectorised -- when applied on a vector of length $n$,
they yield a vector of length $n$ in result.

```{r}
sqrt(1:9)
```

---

Also note the following operations on *logical* vectors:

```{r}
z <- 1:10
all(z >= 5) # are all values TRUE?
any(z >= 5) # is there any value TRUE?
sum(z >= 5) # how many TRUE values are there?
mean(z >= 5) # what is the proportion of TRUE values?
```

The behaviour of `sum()` and `mean()` is dictated by the fact
that, when interpreted in numeric terms, `TRUE==1`  and `FALSE==0`.

### Matrices

---

Vectors are 1-dimensional objects -- they represent sequences of values.

Matrices are 2-dimensional objects -- they represent tabular data.

```{r}
A <- matrix(c(1, 2, 3, 4, 5, 6), byrow=TRUE, nrow=2)
dim(A) # number of rows, number of columns
A
```

Using mathematical notation, above we have defined $\mathbf{A}\in\mathbb{R}^{2\times 3}$:

\[
\mathbf{A}=
\left[
\begin{array}{cccc}
a_{1,1} & a_{1,2} & a_{1,3} \\
a_{2,1} & a_{2,2} & a_{2,3} \\
\end{array}
\right]
=
\left[
\begin{array}{cccc}
1 & 2 & 3 \\
4 & 5 & 6 \\
\end{array}
\right]
\]

---

Other ways to create a matrix:

```{r}
rbind(1:3, 4:6, 7:9) # row bind
rbind(1:3, 4:6, 7:9) # column bind
```


On a side note, R `data.frame`s are similar to matrices but are used
to store tabular data of potentially different types in each column.

---

Note that the omission of `byrow=TRUE` yields the following default behaviour
of the `matrix()` function:

```{r}
matrix(c(1, 2, 3, 4, 5, 6), nrow=2)
```

In other words, elements are read in a column-major manner.

> (\*) This is exactly how R stores the underlying data in RAM.


---

Also take notice of the fact that "flat" vectors are promoted
to column vectors, i.e., matrices with one column:

```{r}
as.matrix(1:3)
```

---

$\mathbf{A}^T$ denotes the matrix *transpose*:

```{r}
t(A)
```

Hence, $\mathbf{B}=\mathbf{A}^T$ is a matrix such that $b_{i,j}=a_{j,i}$.

In other words, in the transposed matrix, rows become columns and columns become rows.

### Matrix-Scalar Operations

---

Operations such as $s\mathbf{A}$ (multiplication of a matrix
by a scalar), $-\mathbf{A}$, $s+\mathbf{A}$ etc.
are applied on each element of the input matrix:

```{r}
(-1)*A
```


### Matrix-Matrix Operations

---



If $\mathbf{A},\mathbf{B}\in\mathbb{R}^{n\times p}$
are two matrices of the same sizes, then
$\mathbf{A}+\mathbf{B}$ and
$\mathbf{A}-\mathbf{B}$ are understood elementwise,
i.e., they result in $\mathbf{C}\in\mathbb{R}^{n\times p}$
such that $c_{i,j}=a_{i,j}\pm b_{i,j}$.

```{r}
A-A
```

---

In R (but not when we use mathematical notation),
all other arithmetic, logical and comparison operators are also
applied in an elementwise fashion.

```{r}
A*A
(A>2) & (A<=5)
```


### Matrix Multiplication (\*)

---

Mathematically, $\mathbf{A}\mathbf{B}$
denotes the **matrix multiplication**. It is a very different operation
to the elementwise multiplication.


```{r}
(A <- rbind(c(1, 2), c(3, 4)))
(I <- rbind(c(1, 0), c(0, 1)))
A %*% I # matrix multiplication
```

This is not the same as the elementwise `A*I`.

---

Matrix multiplication can only be performed on two matrices of
*compatible sizes* -- the number of columns in the left matrix must match
the number of rows in the right operand.

Given $\mathbf{A}\in\mathbb{R}^{n\times p}$
and $\mathbf{B}\in\mathbb{R}^{p\times m}$, their multiply is a matrix
$\mathbf{C}=\mathbf{A}\mathbf{B}\in\mathbb{R}^{n\times m}$
such that $c_{i,j}$ is the dot product of the $i$-th row in $\mathbf{A}$
and the $j$-th column in $\mathbf{B}$:
\[
c_{i,j} = \mathbf{a}_{i,\cdot} \cdot \mathbf{b}_{\cdot,j}
= \sum_{k=1}^p a_{i,k} b_{k, j}
\]
for $i=1,\dots,n$ and $j=1,\dots,m$.

---

> (\*) Note that $\mathbf{A}^T \mathbf{A}$
gives the matrix that consists of the dot products of all the pairs
of columns in $\mathbf{A}$.

```{r}
crossprod(A) # same as t(A) %*% A
```

In the next chapter we will learn about the Pearson linear
correlation coefficient which can be beautifully expressed this way.


> (\*) As an exercise, I recommend that you multiply
a few simple matrices of sizes $2\times 2$, $2\times 3$, $3\times 2$ etc.
using pen and paper and check the results in R. This will become easy
once you get some practice.

<!-- TODO: graphical interpretation -->


---

Also remember that, mathematically,
*squaring* a matrix is done in terms of matrix multiplication,
i.e., $\mathbf{A}^2 = \mathbf{A}\mathbf{A}$.

It can only be performed on *square* matrices, i.e., ones with the same number
of rows and columns.

This is again different than R's elementwise `A^2`.


<!-- TODO? Matrix inverse?? Do we need it? we might for linear regression, but it's already too much -->

<!-- TODO? Matrix ops and the recycling rule -->


### Matrix-Vector Operations

---

Mathematically, there is no generally agreed upon
convention defining arithmetic operations between matrices and vectors.

(\*) The only exception is the  matrix -- vector multiplication in the case
where a vector is a column or a row vector, i.e., in fact, a matrix.

Hence, given $\mathbf{A}\in\mathbb{R}^{n\times p}$ we may write
$\mathbf{A}\mathbf{x}$
only if $\mathbf{x}\in\mathbb{R}^{p\times 1}$ is a column vector.

Similarly, $\mathbf{y}\mathbf{A}$ makes only sense
whenever $\mathbf{y}\in\mathbb{R}^{1\times n}$ is a row vector.

> Please take notice of the fact that we consistently
discriminate between different bold math fonts and letter cases:
$\mathbf{X}$ is a matrix, $\mathbf{x}$ is a row or column vector
(still a matrix, but a sequence-like one)
and $\boldsymbol{x}$ is an ordinary vector (1-dimensional sequence).

> This is why, e.g., the $i$-th row of $\mathbf{X}$ is denoted
with $\mathbf{x}_{i,\cdot}$ -- to emphasise that it is a row vector.

<!-- in R an object of type vector gets promoted to a column vector, i.e., a matrix with 1 column -->



---

However, in R, we might sometimes wish to vectorise
an arithmetic operation between a matrix and a vector in a row- or column-wise
fashion.

For example, if $\mathbf{A}\in\mathbb{R}^{n\times p}$
and $\mathbf{m}\in\mathbb{R}^{1\times p}$ is a row vector,
we might want to subtract $m_i$ from each element in the $i$-th column.

Here, the `apply()` function comes in handy:

* `apply(A, 1, f)` applies a given function $f$ on each *row* of $\mathbf{A}$.
* `apply(A, 2, f)` applies a given function $f$ on each *column* of $\mathbf{A}$.

Usually, either $f$ returns a single value (when we wish to aggregate
all the elements in a row/column) or returns the same number of values
(when we wish to transform a row/column).

---

Example: to create a *centred* version of a given matrix,
we need to subtract from each element the arithmetic mean of its column.

```{r}
(A <- cbind(c(1, 2), c(2, 4), c(5, 8)))
(m <- apply(A, 2, mean)) # same as colMeans(A)
t(apply(A, 1, function(r) r-m)) # note the transpose here
```

---

The above is equivalent to:

```{r}
apply(A, 2, function(c) c-mean(c))
```



## Outro

### Remarks

---

In supervised learning, with each input point,
there's an associated reference output value.

Learning a model = constructing a function that approximates
(minimising some error measure) the given data.

Regression = the output variable $Y$ is continuous.

We studied linear models with a single independent variable based on
the least squares (SSR) fit.

In the next part we will extend this setting to the case
of many variables, i.e., $p>1$, called multiple regression.




### Further Reading

#### {.allowframebreaks .unnumbered}

Recommended further reading:

- [@islr: Chapters 1, 2 and 3]

Other:

- [@esl: Chapter 1, Sections 3.2 and 3.3]
