# Classification with K-Nearest Neighbours

<!-- (C) 2020 Marek Gagolewski, https://www.gagolewski.com -->



## Introduction

### Classification Task

---

Let $\mathbf{X}\in\mathbb{R}^{n\times p}$ be an input matrix
that consists of $n$ points in a $p$-dimensional space (each of the $n$ objects
is described by means of $p$ numerical features).

Recall that in supervised learning, with each
$\mathbf{x}_{i,\cdot}$ we associate the desired output $y_i$.

\[
\mathbf{X}=
\left[
\begin{array}{cccc}
x_{1,1} & x_{1,2} & \cdots & x_{1,p} \\
x_{2,1} & x_{2,2} & \cdots & x_{2,p} \\
\vdots & \vdots & \ddots & \vdots \\
x_{n,1} & x_{n,2} & \cdots & x_{n,p} \\
\end{array}
\right]
\qquad
\mathbf{y} = \left[
\begin{array}{c}
{y}_{1} \\
{y}_{2} \\
\vdots\\
{y}_{n} \\
\end{array}
\right].
\]


. . .



In this chapter we are  interested in  **classification** tasks;
we assume that each $y_i$ is a *label*  (e.g., a character string) --
it is of quantitative/categorical type.

---

Most commonly, we are faced with **binary classification** tasks
where there are only two possible distinct labels.

We traditionally denote them with $0$s and $1$s.

For example:

0       | 1
--------|------------
no      | yes
false   | true
failure | success
healthy | ill


On the other hand, in **multiclass classification**,
we assume that each $y_i$ takes more than two possible values.

---

Example plot of a synthetic dataset
with the reference binary $y$s.

The "true" decision boundary is at $X_1=0$ but the classes
slightly overlap (dataset is a bit noisy).



```{r,echo=FALSE,fig.height=5}
par(mar=c(4,4,0.5,0.5))
set.seed(123)
n0 <- 50 # n0 points in class 0
n1 <- 50 # n1 points in class 1
Xs <- rbind(
    cbind(rnorm(n0, -1, 1), rnorm(n0, 0, 1)), # N( (-1, 0), (1, 1) )
    cbind(rnorm(n1, +1, 1), rnorm(n1, 0, 1))  # N( (+1, 0), (1, 1) )
)
Ys <- factor(rep(c("0", "1"), c(n0, n1)))

plot(Xs[,1], Xs[,2], col=Ys, pch=c(1,2)[Ys], las=1, xlab="X1", ylab="X2", asp=1)
legend("topleft", col=c(1,2), pch=c(1,2), legend=c("Class 0", "Class 1"))
abline(v=0, lty=3, col="blue")
```


### Data

---

For illustration, let's consider the Wine Quality dataset
that can be downloaded from the UCI Machine Learning Repository
(https://archive.ics.uci.edu/ml/datasets/Wine+Quality) --
white wines only.


```{r input1}
wines <- read.csv("datasets/winequality-all.csv", comment="#")
wines <- wines[wines$color == "white",]
(n <- nrow(wines)) # number of samples
```

These are Vinho Verde wine samples from the north of Portugal,
see https://www.vinhoverde.pt/en/homepage.

There are 11 physicochemical features reported.
Moreover, there is a wine rating (which we won't consider here)
on the scale 0 (bad) to 10 (excellent)
given by wine experts.

Source: P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.
Modeling wine preferences by data mining from physicochemical properties. *Decision Support Systems*, **47**(4):547-553, 2009.


---

The input matrix $\mathbf{X}\in\mathbb{R}^{n\times p}$
consists of the first 10 numeric variables:

```{r input2,dependson='input1'}
X <- as.matrix(wines[,1:10])
dim(X)
head(X, 2) # first two rows
```




---

The 11th variable measures the amount of alcohol (in %).

We will convert this dependent variable to a binary one:

- 0 == (`alcohol  < 12`) == lower-alcohol wines1
- 1 == (`alcohol >= 12`) == higher-alcohol wines

```{r input3,dependson='input2'}
# recall that TRUE == 1
Y <- factor(as.character(as.numeric(wines$alcohol >= 12)))
table(Y)
```


Now $(\mathbf{X},\mathbf{y})$ is a basis for an interesting (yet challenging)
binary classification task.





### Training and Test Sets

---

Recall that we are genuinely  interested in the construction of supervised learning models for the two following purposes:

- **description** -- to explain a given dataset in simpler terms,
- **prediction** -- to forecast the values of the dependent variable
for inputs that are yet to be observed.

In the latter case:

- we don't want our models to  *overfit* to current data,
- we want our models to *generalise* well
to new data.

. . .

One way to assess if a model has sufficient predictive power is based
on a random **train-test split** of the original dataset:

- *training sample*  (usually 60-80% of the observations) -- used to construct a model,
- *test sample* (remaining 40-20%) -- used to assess the goodness of fit.

> **Test sample must not be used in the training phase!** (No cheating!)


---

60/40% train-test split in R:

```{R input4,dependson='input3'}
set.seed(123) # reproducibility matters
random_indices <- sample(n)
head(random_indices) # preview
# first 60% of the indices (they are arranged randomly)
# will constitute the train sample:
train_indices <- random_indices[1:floor(n*0.6)]
X_train <- X[train_indices,]
Y_train <- Y[train_indices]
# the remaining indices (40%) go to the test sample:
X_test  <- X[-train_indices,]
Y_test  <- Y[-train_indices]
```





### Discussed Methods

---

Our aim is to build a classifier that takes 10 wine physicochemical
features and determines whether it's a "strong" wine.


We will discuss 3 simple and educational (yet practically useful)
classification algorithms:

- *K-nearest neighbour scheme* -- this chapter,
- *Decision trees* -- the next chapter,
- *Logistic regression* -- the next chapter.




## K-nearest Neighbour Classifier

### Introduction

---

> "If you don't know what to do in a situation, just act like the people around you"

For some integer $K\ge 1$, the **K-Nearest Neighbour  (*K-NN*) Classifier**
proceeds as follows.

To classify a new point $\mathbf{x}'$:

1. find the $K$ nearest neighbours of a given point $\mathbf{x}'$ amongst the points in the train set,
denoted $\mathbf{x}_{i_1,\cdot}, \dots, \mathbf{x}_{i_K,\cdot}$:
    a. compute the Euclidean distances between $\mathbf{x}'$ and each $\mathbf{x}_{i,\cdot}$ from the train set,
    \[d_i = \|\mathbf{x}'-\mathbf{x}_{i,\cdot}\|\]
    b. order $d_i$s in increasing order,
    $d_{i_1} \le d_{i_2} \le \dots \le d_{i_K}$
    c. pick first $K$ indices (these are the *nearest* neighbours)
2. fetch the corresponding reference labels $y_{i_1}, \dots, y_{i_K}$
3. return their *mode* as a result, i.e., the most frequently occurring label (a.k.a. *majority vote*)

<!-- > If a mode is not unique, return a randomly chosen mode (ties are broken at random). -->

---







Here is how $K$-NN classifier works on a synthetic 2D dataset.

Gray and pink regions depict how new points would be classified.

In particular 1-NN is "greedy" in the sense that we just
locate the nearest point.




```{r,echo=FALSE,plot_knn,cache=TRUE}
plot_knn <- function(K) {
    xx1 <- seq(-4, 4, length.out=250)
    xx2 <- seq(-4, 4, length.out=250)
    xx <- expand.grid(xx1, xx2)
    yy <- FNN::knn(Xs, xx, Ys, k=K)
    par(mar=c(4,4,2.5,0.5))
    image(xx1, xx2, matrix(as.numeric(yy)-1, nrow=length(xx1), ncol=length(xx2)),
          col=c("#00000044", "#ff000044"), las=1, xlab="X1", ylab="X2", asp=1,
          main=sprintf("K-NN Class Bounds, K=%d", K))


    points(Xs[,1], Xs[,2], col=Ys, pch=c(1,2)[Ys])
    legend("topleft", col=c(1,2), pch=c(1,2), legend=c("Class 0", "Class 1"), bg="white")
    abline(v=0, lty=3, col="blue")
}
```



```{r,echo=FALSE,fig.height=5,cache=TRUE,dependson='plot_knn'}
plot_knn(1)
```

---


```{r,echo=FALSE,fig.height=5,cache=TRUE,dependson='plot_knn'}
plot_knn(3)
```

Increasing $K$ somehow smoothens the decision boundary (this makes it
less "local" and more "global").

---


```{r,echo=FALSE,fig.height=5,cache=TRUE,dependson='plot_knn'}
plot_knn(25)
```


Recall that the "true" decision boundary for this synthetic dataset
is at $X_1=0$. The 25-NN classifier did quite a good job.


### Example in R

---



We shall be calling the `knn()` function from package `FNN`
to classify the points from the test sample
extracted from the `wines` dataset:

```{r,message=FALSE}
library("FNN")
```


Let us make prediction using the 5-nn classifier:

```{r}
Y_knn5 <- knn(X_train, X_test, Y_train, k=5)
head(Y_test, 28) # True Ys
head(Y_knn5, 28) # Predicted Ys
mean(Y_test == Y_knn5) # accuracy
```


---


9-nn classifier:

```{r}
Y_knn9 <- knn(X_train, X_test, Y_train, k=9)
head(Y_test, 28) # True Ys
head(Y_knn9, 28) # Predicted Ys
mean(Y_test == Y_knn9) # accuracy
```



### Feature Engineering

---

Note that the Euclidean distance that we used above
implicitly assumes that every feature (independent variable)
is on the same scale.

However, when dealing with, e.g., physical quantities,
we often perform conversions of units of measurement (kg → g, feet → m etc.).

Transforming a single feature may drastically change the metric structure of the dataset
and therefore highly affect the obtained predictions.

---

To "bring data to the same scale", we often apply a trick called **standardization**.

Computing the so-called **Z-scores** of the $j$-th feature, $\mathbf{x}_{\cdot,j}$,
is done by subtracting from each observation the sample mean and dividing the result by the sample
standard deviation:

\[z_{i,j} = \frac{x_{i,j}-\bar{x}_{\cdot,j}}{s_{x_{\cdot,j}}}\]

This a new feature $\mathbf{z}_{\cdot,j}$ that always has mean 0 and standard deviation of 1.

Moreover, it is *unit-less* (e.g., we divide a value in kgs by a value in kgs,
the units are cancelled out).


Z-scores are easy to interpret, e.g., 0.5 denotes an observation
that is 0.5 standard deviations above the mean.

---

Let us compute `Z_train` and `Z_test`, being the standardised versions of `X_train`
and `X_test`, respectively.

```{r}
means <- apply(X, 2, mean) # column means
sds   <- apply(X, 2, sd)   # column standard deviations
Z_train <- X_train # to be done
Z_test  <- X_test  # to be done
for (j in 1:ncol(X)) {
    Z_train[,j] <- (Z_train[,j]-means[j])/sds[j]
    Z_test[,j]  <- (Z_test[,j] -means[j])/sds[j]
}
```


Alternatively:

```{r}
Z_train <- t(apply(X_train, 1, function(c) (c-means)/sds))
Z_test  <- t(apply(X_test,  1, function(c) (c-means)/sds))
```


---

Let us compute the accuracy of K-NN classifiers acting on standardised data.

```{r,message=FALSE}
Y_knn5s <- knn(Z_train, Z_test, Y_train, k=5)
mean(Y_test == Y_knn5s) # accuracy
Y_knn9s <- knn(Z_train, Z_test, Y_train, k=9)
mean(Y_test == Y_knn9s) # accuracy
```

The accuracy is much better.


---

Standardisation is an example of *feature engineering*.

Good models rarely work well "straight out of the box" -- if that was the case,
we wouldn't need data scientists and machine learning engineers!

To increase models' accuracy, we often spend a lot of time:

* cleansing data (e.g., removing outliers)
* extracting new features
* transforming existing features
* trying to find a set of features that are relevant

This is the "more art than science" part of data science (sic!), and
hence most textbooks are not really eager for discussing such topics
(including this one).

Sorry, this is sad but true. The solutions that work well in the case of dataset
A may fail in the B case and vice versa. However, the more exercises you solve,
the greater the arsenal of ideas/possible approaches you will have at hand
when dealing with real-world problems.

---

Feature selection -- example (manually selected columns):

```{r,message=FALSE}
features <- c("density", "residual.sugar")
Y_knn5s <- knn(Z_train[,features], Z_test[,features],
    Y_train, k=5)
mean(Y_test == Y_knn5s) # accuracy
Y_knn9s <- knn(Z_train[,features], Z_test[,features],
    Y_train, k=9)
mean(Y_test == Y_knn9s) # accuracy
```

> Try to find a combination of 2-4 features (by guessing or applying magic tricks)
that increases the accuracy of a $K$-NN classifier on this dataset.




## Model Assessment and Selection

### Performance Metrics

---


Recall that $y_i$ denotes the true label associated with the $i$-th observation.

Let $\hat{y}_i$ denote the classifier's output for a given $\mathbf{x}_{i,\cdot}$.

Ideally, we'd wish that $\hat{y}_i=y_i$.

Sadly, in practice we will make errors.

Here are the 4 possible situations (true vs. predicted label):


.              | $y_i=0$                        | $y_i=1$
---------------|--------------------------------|-------------
$\hat{y}_i=0$  | **True Negative**              | False Negative (Type II error)
$\hat{y}_i=1$  | False Positive (Type I error)  | **True Positive**

Note that the terms **positive** and **negative** refer to
the classifier's output, i.e., occur when $\hat{y}_i$ is equal to $1$ and $0$, respectively.


---

A **confusion matrix** is used to summarise
the correctness of predictions for the whole sample:


```{r}
Y_pred <- knn(Z_train, Z_test, Y_train, k=9)
(C <- table(Y_pred, Y_test))
```

For example,

```{r}
C[1,1] # number of TNs
C[2,1] # number of FPs
```


---


**Accuracy** is the ratio of the correctly classified instances
to all the instances.

In other words, it is the probability of making a correct prediction.

\[
\text{Accuracy} = \frac{\text{TP}+\text{TN}}{\text{TP}+\text{TN}+\text{FP}+\text{FN}}
= \frac{1}{n} \sum_{i=1}^n \mathbb{I}\left(
y_i = \hat{y}_i
\right)
\]
where $\mathbb{I}$ is the indicator function,
$\mathbb{I}(l)=1$ if logical condition $l$ is true and $0$ otherwise.

```{r}
mean(Y_test == Y_pred) # accuracy
(C[1,1]+C[2,2])/sum(C) # equivalently
```

---

In many applications we are dealing with **unbalanced problems**, where
the case $y_i=1$ is relatively rare,
yet predicting it  correctly is much more important than being
accurate with respect to class $0$.

> Think of medical applications, e.g., HIV testing
or tumour diagnosis.

In such a case, *accuracy* as a metric fails to quantify what we are aiming for.

> If only 1% of the cases have true $y_i=1$,
then a dummy classifier that always
outputs $\hat{y}_i=0$ has 99% accuracy.

Metrics such as precision and recall (and their aggregated version, F-measure)
aim to address this problem.





---

**Precision**



\[
\text{Precision} = \frac{\text{TP}}{\text{TP}+\text{FP}}
\]


If the classifier outputs $1$,
what is the probability that this is indeed true?

```{r}
C[2,2]/(C[2,2]+C[2,1]) # Precision
```

---

**Recall** (a.k.a. sensitivity, hit rate or true positive rate)

\[
\text{Recall} = \frac{\text{TP}}{\text{TP}+\text{FN}}
\]

If the true class is $1$, what is the probability that the classifier
will detect it?


```{r}
C[2,2]/(C[2,2]+C[1,2]) # Recall
```


---

> Precision or recall? It depends on an application.
> Think of medical diagnosis, medical screening, plagiarism detection, etc.
> --- which measure is more important in each of the settings listed?




As a compromise, we can use the **F-measure**
(a.k.a. $F_1$-measure),
which is the harmonic mean of precision
and recall:

\[
\text{F} = \frac{1}{
    \frac{
        \frac{1}{\text{Precision}}+\frac{1}{\text{Recall}}
    }{2}
}
=
\left(
\frac{1}{2}
\left(
\text{Precision}^{-1}+\text{Recall}^{-1}
\right)
\right)^{-1}
=
\frac{\text{TP}}{\text{TP} + \frac{\text{FP} + \text{FN}}{2}}
\]


> Show that the above equality holds.

```{r}
C[2,2]/(C[2,2]+0.5*C[1,2]+0.5*C[2,1]) # F
```


---

The following function can come in handy in the future:

```{r}
get_metrics <- function(Y_pred, Y_test)
{
    C <- table(Y_pred, Y_test) # confusion matrix
    stopifnot(dim(C) == c(2, 2))
    c(Acc=(C[1,1]+C[2,2])/sum(C), # accuracy
      Prec=C[2,2]/(C[2,2]+C[2,1]), # precision
      Rec=C[2,2]/(C[2,2]+C[1,2]), # recall
      F=C[2,2]/(C[2,2]+0.5*C[1,2]+0.5*C[2,1]), # F-measure
      # Confusion matrix items:
      TN=C[1,1], FN=C[1,2],
      FP=C[2,1], TP=C[2,2]
    ) # return a named vector
}
```


---


```{r}
get_metrics(Y_pred, Y_test)
```


### How to Choose K for K-NN Classification?

---

We haven't yet considered the question which $K$ yields *the best*
classifier.

Best == one that has the highest *predictive power*.

Best == with respect to some chosen metric (accuracy, recall, precision, F-measure, ...)

Let us study how the metrics on the test set change as functions of the number of nearest neighbours considered, $K$.

---


Auxiliary function:

```{r}
knn_metrics <- function(k, X_train, X_test, Y_train, Y_test)
{
    Y_pred <- knn(X_train, X_test, Y_train, k=k) # classify
    get_metrics(Y_pred, Y_test)
}
```

For example:

```{r,cache=TRUE}
knn_metrics(5, Z_train, Z_test, Y_train, Y_test)
```


---


Example call to evaluate metrics as a function of different $K$s:

```{r,cache=TRUE}
Ks <- seq(1, 19, by=2)
Ps <- as.data.frame(t(
    sapply(Ks, # on each element in this vector
        knn_metrics,     # apply this function
        Z_train, Z_test, Y_train, Y_test # aux args
    )))
```


> Note that `sapply(X, f, arg1, arg2, ...)`
outputs a list `Y` such that
`Y[[i]] = f(X[i], arg1, arg2, ...)`
which is then simplified to a matrix.

> We transpose this result, `t()`, in order to get each metric
corresponding to different columns in the result.

> As usual, if you keep wondering, e.g.,  why `t()`, play with the code yourself -- it's fun fun fun.


---


Example results:

```{r}
round(cbind(K=Ks, Ps), 2)
```

A picture is worth a thousand tables though (see `?matplot` in R).



---


```{r,echo=FALSE,cache=TRUE,fig.height=6}
Ks <- seq(1, 51, by=2)
Ps <- as.data.frame(t(sapply(Ks, # on each element in this vector
  knn_metrics,     # apply this function
  Z_train, Z_test, Y_train, Y_test # also passing these as args
)))
par(mar=c(4,4,2.5,0.5))
matplot(Ks, Ps[,1:4], las=1, xlab="K", ylab="Metric", type="l", ylim=c(0,1), main="Standardised Data (Z_train, Z_test)")
legend("top", legend=names(Ps[,1:4]), col=1:4, lty=1:4, ncol=4)
```




---


```{r,echo=FALSE,cache=TRUE,fig.height=6}
Ks <- seq(1, 51, by=2)
Ps <- as.data.frame(t(sapply(Ks, # on each element in this vector
  knn_metrics,     # apply this function
  X_train, X_test, Y_train, Y_test # also passing these as args
)))
par(mar=c(4,4,2.5,0.5))
matplot(Ks, Ps[,1:4], las=1, xlab="K", ylab="Metric", type="l", ylim=c(0,1), main="Raw Data (X_train, X_test)")
legend("top", legend=names(Ps[,1:4]), col=1:4, lty=1:4, ncol=4)
```

<!--

(\*) **ROC** (Receiver Operating Characteristic) curve:

```{r,fig.height=5}
TPR <- Ps$TP/(Ps$TP+Ps$FN) # True Positive Rate (recall)
FPR <- Ps$FP/(Ps$FP+Ps$TN) # False Positive Rate
plot(FPR, TPR, asp=1, xlim=c(0,1), ylim=c(0,1), las=1)
abline(a=0, b=1, lty=3)
```

-->





### Training, Validation and Test sets

---


In the $K$-NN classification task, there are many hyperparameters to tune up:

- Which $K$ should we choose?

- Should we standardise the dataset?

- Which variables should be taken into account when computing the Euclidean distance?

<!--
- Which metric should be used?
-->

> **If we select the best hyperparameter set based on test
sample error, we will run into the trap of overfitting again**.

> This time we'll be overfitting to the test set --- the model that is optimal
for a given test sample doesn't have to generalise well to other test samples (!).

---


In order to overcome this problem,
we can perform a random **train-validation-test split** of the original dataset:

- *training sample*  (e.g., 60%) -- used to construct the models
- *validation sample* (e.g., 20%) -- used to tune the hyperparameters of the classifier
- *test sample* (e.g., 20%) -- used to assess the goodness of fit

> (\*) If our dataset is too small,
we can use various *crossvalidation* techniques
instead of a train-validate-test split.


<!--By the way, this is how most data mining competitions are assessed --
you will never have access to the final test sample used to determine the winner.
The best you can do is to "guess".-->


---

An example way to perform a 60/20/20% train-validation-test split:

```{R}
set.seed(123) # reproducibility matters
random_indices <- sample(n)
n1 <- floor(n*0.6)
n2 <- floor(n*0.8)
X2_train <- X[random_indices[1     :n1], ]
Y2_train <- Y[random_indices[1     :n1]  ]
X2_valid <- X[random_indices[(n1+1):n2], ]
Y2_valid <- Y[random_indices[(n1+1):n2]  ]
X2_test  <- X[random_indices[(n2+1):n ], ]
Y2_test  <- Y[random_indices[(n2+1):n ]  ]
stopifnot(nrow(X2_train)+nrow(X2_valid)+nrow(X2_test)
    == nrow(X))
```


> As an exercise, find the best $K$ on the validation set and compute the error metrics
on the test set.




## Implementing a K-NN Classifier (\*)


### Factor Data Type

---

Recall that (see Appendix B for more details)
`factor` type in R is a very convenient means to encode categorical data
(such as $\mathbf{y}$):

```{r}
x <- c("yes", "no", "no", "yes", "no")
f <- factor(x, levels=c("no", "yes"))
f
table(f) # counts
```

---

Internally, objects of type `factor` are represented as integer vectors
with elements in $\{1,\dots,M\}$, where $M$ is the number of possible levels.


Labels, used to "decipher" the numeric codes, are stored separately.

```{r}
as.numeric(f) # 2nd label, 1st label, 1st label etc.
levels(f)
levels(f) <- c("failure", "success") # re-encode
f
```



### Main Routine (\*)

---

Let us implement a K-NN classifier ourselves
by using a top-bottom approach.

We will start with a general description of the admissible inputs
and the expected output.

Then we will arrange the processing of data into
conveniently manageable chunks.

The function's declaration will look like:

```{r,eval=FALSE}
our_knn <- function(X_train, X_test, Y_train, k=1) {
    # k=1 denotes a parameter with a default value
    # ...
}
```

---

Load an example dataset on which we will test our algorithm:

```{r}
wines <- read.csv("datasets/winequality-all.csv", comment="#")
wines <- wines[wines$color == "white",]
X <- as.matrix(wines[,1:10])
Y <- factor(as.character(as.numeric(wines$alcohol >= 12)))
```


Note that `Y` is now a factor object.


Train-test split:

```{r}
set.seed(123)
random_indices <- sample(n)
train_indices <- random_indices[1:floor(n*0.6)]
X_train <- X[train_indices,]
Y_train <- Y[train_indices]
X_test  <- X[-train_indices,]
Y_test  <- Y[-train_indices]
```


---





First, we should specify the type and form of the arguments
we're expecting:

```{r,eval=FALSE,our_knn_paramchecks}
# this is the body of our_knn() - part 1
stopifnot(is.numeric(X_train), is.matrix(X_train))
stopifnot(is.numeric(X_test), is.matrix(X_test))
stopifnot(is.factor(Y_train))
stopifnot(ncol(X_train) == ncol(X_test))
stopifnot(nrow(X_train) == length(Y_train))
stopifnot(k >= 1)
n_train <- nrow(X_train)
n_test  <- nrow(X_test)
p <- ncol(X_train)
M <- length(levels(Y_train))
```

Therefore,

$\mathtt{X\_train}\in\mathbb{R}^{\mathtt{n\_train}\times \mathtt{p}}$,
$\mathtt{X\_test}\in\mathbb{R}^{\mathtt{n\_test}\times \mathtt{p}}$ and
$\mathtt{Y\_train}\in\{1,\dots,M\}^{\mathtt{n\_train}}$

> Recall that R `factor` objects are internally encoded as integer vectors.

---

Next, we will call the (to-be-done) function `our_get_knnx()`,
which seeks nearest neighbours of all the points:

```{r,eval=FALSE,our_knn_flow1}
# our_get_knnx returns a matrix nn_indices of size n_test*k,
# where nn_indices[i,j] denotes the index of
# X_test[i,]'s j-th nearest neighbour in X_train.
# (It is the point X_train[nn_indices[i,j],]).
nn_indices <- our_get_knnx(X_train, X_test, k)
```


---

Then, for each point in `X_test`,
we fetch the labels corresponding to its nearest neighbours
and compute their mode:

```{r,eval=FALSE,our_knn_flow2}
Y_pred <- numeric(n_test) # vector of length n_test
# For now we will operate on the integer labels in {1,...,M}
Y_train_int <- as.numeric(Y_train)
for (i in 1:n_test) {
    # Get the labels of the NNs of the i-th point:
    nn_labels_i <- Y_train_int[nn_indices[i,]]
    # Compute the mode (majority vote):
    Y_pred[i] <- our_mode(nn_labels_i) # in {1,...,M}
}
```

Finally, we should convert the resulting integer vector
to an object of type `factor`:

```{r,eval=FALSE,our_knn_flow3}
# Convert Y_pred to factor:
return(factor(Y_pred, labels=levels(Y_train)))
```



```{r,echo=FALSE}
our_knn <- function(X_train, X_test, Y_train, k=1) {

    <<our_knn_paramchecks>>
    <<our_knn_flow1>>
    <<our_knn_flow2>>
    <<our_knn_flow3>>
}
```



<!--
**Test-driven development** -- before writing

```{r}
test_our_knn <- function() {
    # ...
}
```


```{r}
test_our_mode <- function() {
    stopifnot(our_mode(c(1, 1, 1, 1)) == 1)
    stopifnot(our_mode(c(2, 2, 2, 2)) == 2)
    stopifnot(our_mode(c(3, 1, 3, 3)) == 3)
    stopifnot(our_mode(c(1, 1, 3, 3, 2)) %in% c(1, 3))
}
```


-->

### Mode

---

To implement the mode, we can use the `tabulate()` function.

> Read the function's man page, see `?tabulate`.

For example:

```{R}
tabulate(c(1, 2, 1, 1, 1, 5, 2))
```

---

There might be multiple modes -- in such a case, we should pick one at random.

For that, we can use the `sample()` function.

> Read the function's man page, see `?sample`.
> Note that its behaviour is different when it's first argument is a vector of length 1.


---

An example implementation:

```{r}
our_mode <- function(Y) {
    # tabulate() will take care of
    # checking the correctness of Y
    t <- tabulate(Y)
    mode_candidates <- which(t == max(t))
    if (length(mode_candidates) == 1) return(mode_candidates)
    else return(sample(mode_candidates, 1))
}
```

---

```{r,echo=-1}
set.seed(7)
our_mode(c(1, 1, 1, 1))
our_mode(c(2, 2, 2, 2))
our_mode(c(3, 1, 3, 3))
our_mode(c(1, 1, 3, 3, 2))
our_mode(c(1, 1, 3, 3, 2))
```




###  NN Search Routines (\*)

---

Last but not least, we should implement the `our_get_knnx()` function.

It is the function responsible for seeking the indices of nearest neighbours.

It turns out this function will  actually constitute the K-NN classifier's performance
bottleneck in case of big data samples.

```{r}
# our_get_knnx returns a matrix nn_indices of size n_test*k,
# where nn_indices[i,j] denotes the index of
# X_test[i,]'s j-th nearest neighbour in X_train.
# (It is the point X_train[nn_indices[i,j],]).
our_get_knnx <- function(X_train, X_test, k) {
    # ...
}
```


---


A naive approach to `our_get_knnx()` relies on computing all pairwise distances,
and sorting them.

```{r}
our_get_knnx <- function(X_train, X_test, k) {
    n_test <- nrow(X_test)
    nn_indices <- matrix(NA_real_, nrow=n_test, ncol=k)
    for (i in 1:n_test) {
        d <- apply(X_train, 1, function(x)
            sqrt(sum((x-X_test[i,])^2)))
        # now d[j] is the distance
        # between X_train[j,] and X_test[i,]
        nn_indices[i,] <- order(d)[1:k]
    }
    nn_indices
}
```



---

A comparison with `FNN:knn()`:

```{r,cache=TRUE}
system.time(Ya <- knn(X_train, X_test, Y_train, k=5))
system.time(Yb <- our_knn(X_train, X_test, Y_train, k=5))
mean(Ya == Yb) # 1.0 on perfect match
```

Both functions return identical results but our implementation is "slightly" slower.

---



`FNN:knn()` is efficiently written in C++, which is a compiled programming language.

R, on the other hand (just like Python and Matlab) is interpreted, therefore
as a rule of thumb we should consider it an order of magnitude slower (see, however, the Julia language).

Let us substitute our naive implementation with the equivalent one,
but written in C++ (available in the `FNN` package).

> (\*) Note that we could write a C++ implementation ourselves,
see the Rcpp package for seamless R and C++ integration.

---

```{r,cache=TRUE}
our_get_knnx <- function(X_train, X_test, k) {
    # this is used by our_knn()
    FNN::get.knnx(X_train, X_test, k, algorithm="brute")$nn.index
}
system.time(Ya <- knn(X_train, X_test, Y_train, k=5))
system.time(Yb <- our_knn(X_train, X_test, Y_train, k=5))
mean(Ya == Yb) # 1.0 on perfect match
```

---

Note that our solution requires $c\cdot n_\text{test}\cdot n_\text{train}\cdot p$
arithmetic operations for some $c>1$.
The overall cost of sorting is at least $d\cdot n_\text{test}\cdot n_\text{train}\cdot\log n_\text{train}$
for some $d>1$.

This does not scale well with both $n_\text{test}$ and $n_\text{train}$
(think -- big data).

. . .

It turns out that there are special **spatial data structures**
-- such as *metric trees* -- that aim to speed up searching for nearest
neighbours in *low-dimensional spaces* (for small $p$).

> (\*) Searching in high-dimensional spaces is hard due to the so-called
curse of dimensionality.


For example, `FNN::get.knnx()` also implements the so-called
kd-trees.

---


```{r,test_speed,cache=TRUE}
library("microbenchmark")
test_speed <- function(n, p, k) {
    A <- matrix(runif(n*p), nrow=n, ncol=p)
    s <- summary(microbenchmark(
        brute=FNN::get.knnx(A, A, k, algorithm="brute"),
        kd_tree=FNN::get.knnx(A, A, k, algorithm="kd_tree"),
        times=3
    ), unit="s")
    # minima of 3 time measurements:
    structure(s$min, names=as.character(s$expr))
}
```


---

```{r,cache=TRUE,dependson='test_speed'}
test_speed(10000, 2, 5)
test_speed(10000, 5, 5)
test_speed(10000, 10, 5)
test_speed(10000, 20, 5)
```









### Different Metrics (\*)

---

The Euclidean distance is just one particular example
of many possible **metrics** (metric == a mathematical term,
above we have used this term in a more relaxed fashion, when referring
to accuracy etc.).

Mathematically, we say that $d$ is a metric on a set $X$
(e.g., $\mathbb{R}^p$), whenever
it is a function $d:X\times X\to [0,\infty]$ such that for all $x,x',x''\in X$:

- $d(x, x') = 0$ if and only if $x=x'$,
- $d(x, x') = d(x', x)$ (it is symmetric)
- $d(x, x'') \le d(x, x') + d(x', x'')$ (it fulfils the triangle inequality)


> (*) Not all the properties are required in all the applications;
sometimes we might need a few additional ones.

We can easily generalise the way we introduced the K-NN method
to have a classifier that is based on a point's neighbourhood
with respect to any metric.

---

Example metrics on $\mathbb{R}^p$:

- **Euclidean**
\[
d_2(\mathbf{x}, \mathbf{x}') = \| \mathbf{x}-\mathbf{x}' \| = \| \mathbf{x}-\mathbf{x}' \|_2 = \sqrt{ \sum_{i=1}^p (x_i-x_i')^2 }
\]
- **Manhattan** (taxicab)
\[
d_1(\mathbf{x}, \mathbf{x}') = \| \mathbf{x}-\mathbf{x}' \|_1 = { \sum_{i=1}^p |x_i-x_i'| }
\]
- **Chebyshev** (maximum)
\[
d_\infty(\mathbf{x}, \mathbf{x}') = \| \mathbf{x}-\mathbf{x}' \|_\infty = \max_{i=1,\dots,p} |x_i-x_i'|
\]

<!--
These are all examples of $L_p$ metrics, $p\ge 1$:
\[
d_p(\mathbf{x}, \mathbf{x}') = \| \mathbf{x}-\mathbf{x}' \|_p = \left( \sum_{i=1}^p |x_i-x_i'|^p \right)^{1/p}
\]
-->

---

We can define metrics on different spaces too.

For example, the **Levenshtein distance** is a popular choice
for comparing character strings (also DNA sequences etc.)

It is an *edit distance* -- it measures the minimal number of
single-character insertions, deletions or substitutions to change
one string into another.


For instance:

```{r}
adist("happy", "nap")
```

This is because we need 1 substitution and 2 deletions,

happy → nappy → napp → nap.


---




See also:

- the Hamming distance for categorical vectors (or strings of equal lengths),
- the Jaccard distance for sets,
- the Kendall tau rank distance for rankings.

Moreover, R package `stringdist` includes implementations
of numerous string metrics.


<!-- Mahalanobis distance -->




## Outro

### Remarks

---

Note that K-NN is suitable for any kind of multiclass classification.

In the next part we will discuss some other noteworthy classifiers:

- *Decision trees*
- *Logistic regression*


They are restricted to binary (0/1) outputs. They will have to be extended
somehow to allow for more classes.



### Side Note: K-NN Regression

---

The K-Nearest Neighbour scheme is intuitively pleasing.

No wonder it has inspired a similar approach for solving a regression task.

In order to make a prediction for a new point $\mathbf{x}'$:

1. find the K-nearest neighbours of  $\mathbf{x}'$ amongst the points in the train set,
denoted $\mathbf{x}_{i_1,\cdot}, \dots, \mathbf{x}_{i_K,\cdot}$,
2. fetch the corresponding reference outputs $y_{i_1}, \dots, y_{i_K}$,
3. return their arithmetic mean as a result,
\[\hat{y}=\frac{1}{K} \sum_{j=1}^K y_{i_j}.\]





---

Recall our modelling of the Credit Rating ($Y$)
as a function of the average Credit Card Balance ($X$)
based on the `ISLR::Credit` data set.



```{r}
library("ISLR") # Credit dataset
Xc <- as.matrix(as.numeric(Credit$Balance[Credit$Balance>0]))
Yc <- as.matrix(as.numeric(Credit$Rating[Credit$Balance>0]))
```



```{r,message=FALSE}
library("FNN") # knn.reg function
x <- as.matrix(seq(min(Xc), max(Xc), length.out=101))
y1  <- knn.reg(Xc, x, Yc, k=1)$pred
y5  <- knn.reg(Xc, x, Yc, k=5)$pred
y25 <- knn.reg(Xc, x, Yc, k=25)$pred
```

---

```{r, fig.height=5,message=FALSE,echo=-1}
par(mar=c(4,4,0.5,0.5))
plot(Xc, Yc, las=1, col="#666666c0",
    xlab="Balance", ylab="Rating")
lines(x, y1,  col=2, lwd=3)
lines(x, y5,  col=3, lwd=3)
lines(x, y25, col=4, lwd=3)
```






### Further Reading

#### {.allowframebreaks .unnumbered}

Recommended further reading:

- [@esl: Section 13.3]
