<!--
kate: indent-width 4; word-wrap-column 74; default-dictionary en_AU
Copyright (C) 2020-2021, Marek Gagolewski <https://www.gagolewski.com>
This material is licensed under the Creative Commons BY-NC-ND 4.0 License.
-->

# Recommender Systems

```{r chapter-header-motd,echo=FALSE,results="asis"}
cat(readLines("chapter-header-motd.md"), sep="\n")
```

## Introduction

*Recommender (recommendation) systems*
aim to predict the rating a *user* would give to an *item*.

For example:

- playlist generators (Spotify, YouTube, Netflix, ...),
- content recommendations (Facebook, Instagram, Twitter, Apple News, ...),
- product recommendations (Amazon, Alibaba, ...).




Implementing recommender systems, according to [@ricci_etal],
might:

- increase the number of items sold,
- increase users' satisfaction,
- increase users' fidelity,
- allow a company to sell more diverse items,
- allow to better understand what users want.

{ BEGIN exercise }
Think of the last time you found some recommendation useful.
{ END exercise }


They can also increase the users' frustration.

{ BEGIN exercise }
Think of the last time you found a recommendation useless and
irritating. What might be the reasons why you have been provided
with such a suggestion?
{ END exercise }

<!--
 ("Why does this
thing keep recommending me those stupid videos? Ah, wait, I remember
I watched one similar piece that a friend sent me on a chat.")
-->






### The Netflix Prize



In 2006 Netflix (back then a DVD rental company) released one of the most famous
benchmark sets for recommender systems, which helped boost the research
on algorithms in this field.

See https://www.kaggle.com/netflix-inc/netflix-prize-data;
data archived at
https://web.archive.org/web/20090925184737/http://archive.ics.uci.edu/ml/datasets/Netflix+Prize
and https://archive.org/details/nf_prize_dataset.tar


The dataset consists of:

- 480,189 users
- 17,770 movies
- 100,480,507 ratings in the training sample:
    - `MovieID`
    - `CustomerID`
    - `Rating` from 1 to 5
    - `Title`
    - `YearOfRelease` from 1890 to 2005
    - `Date` of rating in the range 1998-11-01 to 2005-12-31




The *quiz set* consists of 1,408,342 ratings
and it was used by the competitors to assess the quality of their
algorithms and compute the leaderboard scores.

Root mean squared error (RMSE) of predicted vs. true rankings was chosen as a
performance metric.

The *test set* of 1,408,789 ratings (not make publicly available)
was used to determine the winner.

On 21 September 2009, the grand prize of US\$1,000,000 was given
to the BellKor's Pragmatic Chaos team which improved over
the Netflix's *Cinematch* algorithm  by 10.06%,
achieving the winning RMSE of 0.8567 on the test subset.



### Main Approaches



Current recommender systems are quite complex and use a fusion
of various approaches, also those based on external knowledge bases.

However, we may distinguish at least two core approaches,
see [@ricci_etal] for more:

- *Collaborative Filtering* is based on the assumption that if
two people interact with the same product,
they're likely to have other interests in common as well.

    > John and Mary both like bananas and apples and dislike spinach. John likes
sushi. Mary hasn't tried sushi yet. It seems they might have similar tastes,
so we recommend that Mary should give sushi a try.

- *Content-based Filtering* builds users' profiles that represent information
    about what kind of products they like.

    > We have discovered that John likes fruit but dislikes vegetables.
    An orange is a fruit (an item similar to those he liked in the past)
    with which John is yet to interact. Thus, it is suggested that John should
    give it a try.



Jim Bennett, at that time the vice president of recommendation systems at Netflix
on the idea behind the original Cinematch algorithm (see https://www.technologyreview.com/s/406637/the-1-million-netflix-challenge/
and https://web.archive.org/web/20070821194257/http://www.netflixprize.com/faq):

> First, you collect 100 million user ratings for about 18,000 movies. Take any two movies and find the people who have rated both of them. Then look to see if the people who rate one of the movies highly rate the other one highly, if they liked one and not the other, or if they didn’t like either movie. Based on their ratings, Cinematch sees whether there’s a correlation between those people. Now, do this for all possible pairs of 65,000 movies.



{ BEGIN exercise }
Is the above an example of the collaborative or context-based filtering?
{ END exercise }


<!--
How does Cinematch do it?

Straightforward statistical linear models with a lot of data conditioning. But a real-world system is much more than an algorithm, and Cinematch does a lot more than just optimize for RMSE. After all, we have a website to support. In production we have to worry about system scaling and performance, and we have additional sources to data we can use to guide our recommendations.

Netflix Prize FAQ https://web.archive.org/web/20070821194257/http://www.netflixprize.com/faq-->




### Formalism



Let $\mathcal{U}=\{ U_1,\dots,U_n \}$ denote the set of $n$ users.

Let $\mathcal{I}=\{ I_1,\dots,I_p \}$ denote the set of $p$ items.


Let $\mathbf{R}\in\mathbb{R}^{n\times p}$ be a user-item matrix such that:
\[
r_{u,i}=\left\{
\begin{array}{ll}
r & \text{if the $u$-th user ranked the $i$-th item as $r>0$}\\
0 & \text{if the $u$-th user hasn't interacted with the $i$-th item yet}\\
\end{array}
\right.
\]


Remark.
: Note that `0` is used to denote a missing value (`NA`) here.


In particular, we can assume:

- $r_{u,i}\in\{0,1,\dots,5\}$ (ratings on the scale 1--5 or no interaction)
- $r_{u,i}\in\{0,1\}$ ("Like" or no interaction)


The aim of an recommender system is to predict the rating $\hat{r}_{u,i}$
that the $u$-th user would give to the $i$-th item provided that currently
$r_{u,i}=0$.





## Collaborative Filtering

### Example




   .   | Apple   | Banana  |  Sushi    |  Spinach |  Orange   |
-------|---------|---------|-----------|----------|-----------|
Anne   |   1     |   5     |   5       |          |   1       |
Beth   |   1     |   1     |   5       |    5     |   1       |
John   |   5     |   5     |           |    1     |           |
Kate   |   1     |   1     |   5       |    5     |   1       |
Mark   |   5     |   5     |   1       |    1     |   5       |
Sara   |   ?     |   5     |           |    ?     |   5       |


In **user-based collaborative filtering**, we seek users with similar
preference profiles/rating patters.

> "User A has similar behavioural patterns as user B, so A should suggested
    with what B likes."

In **item-based collaborative filtering**, we seek items with similar (dis)likeability
structure.

> "Users who (dis)liked X also (dis)liked Y".

{ BEGIN exercise }
Will Sara enjoy her spinach? Will Sara enjoy her apple?
{ END exercise }


An example $\mathbf{R}$ matrix in R:

```{r recommender1}
R <- matrix(
    c(
     1, 5, 5, 0, 1,
     1, 1, 5, 5, 1,
     5, 5, 0, 1, 0,
     1, 1, 5, 5, 1,
     5, 5, 1, 1, 5,
     0, 5, 0, 0, 5
    ), byrow=TRUE, nrow=6, ncol=5,
    dimnames=list(
        c("Anne", "Beth", "John", "Kate", "Mark", "Sara"),
        c("Apple", "Banana", "Sushi", "Spinach", "Orange")
    )
)
```



```{r recommender2}
R
```


### Similarity Measures





Assuming $\mathbf{a},\mathbf{b}$ are two sequences of length $k$
(in our setting, $k$ is equal to either $n$ or $p$),
let $S$ be the following similarity measure between two rating vectors:


\[
S(\mathbf{a},\mathbf{b}) = \frac{ \sum_{i=1}^k a_i b_i
}{
\sqrt{ \sum_{i=1}^k a_i^2 }
\sqrt{ \sum_{i=1}^k b_i^2 }
}
\]

```{r recommender3}
cosim <- function(a, b) sum(a*b)/sqrt(sum(a^2)*sum(b^2))
```

We call it the **cosine similarity**.
We have $S(\mathbf{a},\mathbf{b})\in[-1,1]$,
where we get $1$ for two identical elements.
Similarity of 0 is obtained for two unrelated ("orthogonal") vectors.
For nonnegative sequences, negative similarities are not generated.

> (\*) Another frequently considered similarity measure
is a version of the Pearson correlation coefficient that
ignores all $0$-valued observations,
see also the `use` argument to the `cor()` function.



### User-Based Collaborative Filtering



**User-based** approaches involve comparing each user against every other user
(pairwise comparisons of the rows in $\mathbf{R}$). This yields a similarity degree
between the $i$-th and the $j$-th user:

\[
s^U_{i,j} = S(\mathbf{r}_{i,\cdot},\mathbf{r}_{j,\cdot}).
\]



```{r recommender4}
SU <- matrix(0, nrow=nrow(R), ncol=nrow(R),
    dimnames=dimnames(R)[c(1,1)]) # and empty n*n matrix
for (i in 1:nrow(R)) {
    for (j in 1:nrow(R)) {
        SU[i,j] <- cosim(R[i,], R[j,])
    }
}
```



```{r recommender5}
round(SU, 2)
```



In order to obtain the previously unobserved
rating $\hat{r}_{u,i}$ using the user-based approach, we typically
look for the $K$ most similar users and aggregate their corresponding
scores (for some $K\ge 1$).

More formally, let $\{U_{v_1},\dots,U_{v_K}\}\in\mathcal{U}\setminus\{U_u\}$ be the set
of users maximising $s^U_{u, v_1}, \dots, s^U_{u, v_K}$
and having $r_{v_1, i},\dots,r_{v_K, i}>0$.
Then:
\[
\hat{r}_{u,i} = \frac{1}{K} \sum_{\ell=1}^K r_{v_\ell, i}.
\]

Remark.
: The arithmetic mean can be replaced with, e.g.,
the more or a weighted arithmetic mean where weights are proportional to $s^U_{u, v_\ell}$

This is very similar to the $K$-nearest neighbour heuristic!




```{r recommender6}
K <- 2
(sim <- order(SU["Sara",], decreasing=TRUE))
# sim gives the indices of people in decreasing order
# of similarity to Sara:
dimnames(R)[[1]][sim] # the corresponding names
# Remove those who haven't tried Spinach yet (including Sara):
sim <- sim[ R[sim, "Spinach"]>0 ]
dimnames(R)[[1]][sim]
# aggregate the Spinach ratings of the K most similar people:
mean(R[sim[1:K], "Spinach"])
```



### Item-Based Collaborative Filtering



**Item-based** schemes rely on pairwise comparisons between the items
(columns in $\mathbf{R}$). Hence, a similarity degree between the $i$-th and the $j$-th
item is given by:


\[
s^I_{i,j} = S(\mathbf{r}_{\cdot,i},\mathbf{r}_{\cdot,j}).
\]



```{r recommender7}
SI <- matrix(0, nrow=ncol(R), ncol=ncol(R),
    dimnames=dimnames(R)[c(2,2)]) # an empty p*p matrix
for (i in 1:ncol(R)) {
    for (j in 1:ncol(R)) {
        SI[i,j] <- cosim(R[,i], R[,j])
    }
}
```



```{r recommender8}
round(SI, 2)
```




In order to obtain the previously unobserved
rating $\hat{r}_{u,i}$ using the item-based approach, we typically
look for the $K$ most similar items and aggregate their corresponding
scores (for some $K\ge 1$)

More formally, let $\{I_{j_1},\dots,I_{j_K}\}\in\mathcal{I}\setminus\{I_i\}$ be the set
of items maximising $s^I_{i, j_1}, \dots, s^I_{i, j_K}$ and having $r_{u, j_1},\dots,r_{u, j_K}>0$.    Then:

\[
\hat{r}_{u,i} = \frac{1}{K} \sum_{\ell=1}^K r_{u, j_\ell}.
\]

Remark.
: Similarly to the previous case,
the arithmetic mean can be replaced with, e.g.,
the mode or
a weighted arithmetic mean where weights are proportional to $s^I_{i, j_\ell}$.






```{r recommender9}
K <- 2
(sim <- order(SI["Apple",], decreasing=TRUE))
# sim gives the indices of items in decreasing order
# of similarity to Apple:
dimnames(R)[[2]][sim] # the corresponding item types
# Remove these which Sara haven't tried yet (e.g., Apples):
sim <- sim[ R["Sara", sim]>0 ]
dimnames(R)[[2]][sim]
# aggregate Sara's ratings of the K most similar items:
mean(R["Sara", sim[1:K]])
```




## Exercise: The MovieLens Dataset (\*)

### Dataset



Let's make a few recommendations based on the MovieLens-9/2018-Small
dataset available at
https://grouplens.org/datasets/movielens/latest/,
see also https://movielens.org/ and [@movielens].

The dataset consists of
ca. 100,000 ratings to 9,000 movies by 600 users. It was last updated
on September 2018.

This is already a pretty large dataset! We might run into problems
with memory usage and high run-time.

<!--
The following examples are a bit more difficult to follow
(programming-wise), therefore
we mark them with (\*).
-->








```{r read_movies,cache=TRUE}
movies <- read.csv("datasets/ml-9-2018-small/movies.csv",
    comment.char="#")
head(movies, 4)
nrow(movies)
```




```{r read_ratings,cache=TRUE,dependson='read_movies'}
ratings <- read.csv("datasets/ml-9-2018-small/ratings.csv",
    comment.char="#")
head(ratings, 4)
nrow(ratings)
table(ratings$rating)
```


### Data Cleansing



`movieId`s should be re-encoded, as not every film is mentioned/rated in the database.
We will re-map the `movieId`s to consecutive integers.

```{r cleanse_movies,cache=TRUE,rependson='read_ratings'}
# the list of all rated movieIds:
movieId2 <- unique(ratings$movieId)
# max user Id (these could've been cleaned up too):
(n <- max(ratings$userId))
# number of unique movies:
(p <- length(movieId2))
# remove unrated movies:
movies <- movies[movies$movieId %in% movieId2, ]
```



```{r cleanse_movies2,cache=TRUE,rependson='cleanse_movies'}
# we'll map movieId2[i] to i for each i=1,...,p:
movies$movieId  <- match(movies$movieId, movieId2)
ratings$movieId <- match(ratings$movieId, movieId2)
# order the movies by the new movieId so that
# the movie with Id==i is in the i-th row:
movies <- movies[order(movies$movieId),]
stopifnot(all(movies$movieId == 1:p)) # sanity check
```






We will use a sparse matrix data type (from R package `Matrix`)
to store the ratings data, $\mathbf{R}\in\mathbb{R}^{n\times p}$.

Remark.
: *Sparse* matrices contain many zeros. Instead of storing all the
$np=`r n*p`$ elements, only the lists of non-zero ones are saved,
$`r nrow(ratings)`$ values in total.
This way, we might save a lot of memory.
The drawback is that, amongst others, random access to the elements
in a sparse matrix takes more time.



```{r movielens1,cache=TRUE,dependson='cleanse_movies2'}
library("Matrix")
R <- Matrix(0.0, sparse=TRUE, nrow=n, ncol=p)
# This is a vectorised operation;
# it is faster than a for loop over each row
# in the ratings matrix:
R[cbind(ratings$userId, ratings$movieId)] <- ratings$rating
```


<!--# Not every movie is rated - removing:
R <- R[,apply(R,2,sum)>0]
# Not each user gave a rating - removing:
R <- R[apply(R,1,sum)>0,]-->



Let's preview a few first rows and columns:

```{r movielens2,cache=TRUE,dependson='movielens1'}
R[1:6, 1:18]
```


### Item-Item Similarities



To recall, the cosine similarity between
$\mathbf{r}_{\cdot,i},\mathbf{r}_{\cdot,j}\in\mathbb{R}^n$
is given by:

\[
s_{i,j}^I = S_C(\mathbf{r}_{\cdot,i},\mathbf{r}_{\cdot,j}) = \frac{\sum_{k=1}^n r_{k,i} \, r_{k,j}}{
    \sqrt{\sum_{k=1}^n r_{k,i}^2}\sqrt{\sum_{k=1}^n r_{k,j}^2}
}
\]

In vector/matrix algebra notation, this is:

\[
s_{i,j}^I = S_C(\mathbf{r}_{\cdot,i},\mathbf{r}_{\cdot,j}) = \frac{\mathbf{r}_{\cdot,i}^T\, \mathbf{r}_{\cdot,j}}{
\sqrt{{\mathbf{r}_{\cdot,i}^T\, \mathbf{r}_{\cdot,i}}} \sqrt{{\mathbf{r}_{\cdot,j}^T\, \mathbf{r}_{\cdot,j}}}
}
\]

As $\mathbf{R}\in\mathbb{R}^{n\times p}$,
we can "almost" compute all the $p\times p$ cosine similarities
at once by applying:

\[
\mathbf{S}^I = \frac{\mathbf{R}^T \mathbf{R}}{
\dots
}
\]



Cosine similarities for item-item comparisons:

```{r movielens3,dependson='movielens2',cache=TRUE}
norms <- as.matrix(sqrt(colSums(R^2)))
Rx <- as.matrix(crossprod(R, R))
SI <- Rx/tcrossprod(norms)
SI[is.nan(SI)] <- 0 # there were some divisions by zero
```

Remark.

: `crossprod(A,B)` gives $\mathbf{A}^T \mathbf{B}$
and `tcrossprod(A,B)` gives $\mathbf{A} \mathbf{B}^T$.


### Example Recommendations



```{r movielens4,dependson='movielens3',cache=TRUE}
recommend <- function(i, K, SI, movies) {
    # get K most similar movies to the i-th one
    ms <- order(SI[i,], decreasing=TRUE)
    data.frame(
        Title=movies$title[ms[1:K]],
        SIC=SI[i,ms[1:K]]
    )
}
```




```{r movielens5,dependson='movielens4',cache=TRUE}
movies$title[1215]
recommend(1215, 10, SI, movies)
```



```{r movielens6,dependson='movielens5',cache=TRUE}
movies$title[1]
recommend(1, 10, SI, movies)
```

...and so on.


### Clustering



All our ratings are $r_{i,j}\ge 0$, therefore the cosine similarity is
$s_{i,j}^I\ge 0$. Moreover, it holds $s_{i,j}^I\le 1$.
Thus, a cosine similarity matrix can be turned into
a dissimilarity matrix:

```{r movielens7,dependson='movielens6',cache=TRUE}
DI <- 1.0-SI
DI[DI<0] <- 0.0 # account for numeric inaccuracies
DI <- as.dist(DI)
```




This enables us to perform, e.g., the cluster analysis of items:

```{r movielens8,dependson='movielens7',cache=TRUE,fig.cap="Cluster dendrogram for the movies"}
library("genie")
h <- hclust2(DI)
plot(h, labels=FALSE, ann=FALSE); box()
```


A 14-partition might look nice, let's give it a try:

```{r movielens9,dependson='movielens8',cache=TRUE}
c <- cutree(h, k=14)
```




Example movies in the 3rd cluster:

`r paste(head(movies$title[c==3], 20), collapse=", ")`

The definitely have something in common!




Example movies in the 1st cluster:

`r paste(head(movies$title[c==1], 20), collapse=", ")`


... and so forth.



## Outro

### Remarks



Good recommender systems are perfect tools to increase the revenue
of any user-centric enterprise.

Not a single algorithm, but an ensemble (a proper combination) of different approaches
is often used in practice, see the Further Reading section below for the detailed
information of the Netflix Prize winners.

Recommender systems are an interesting fusion of the techniques we
have already studied -- linear models, K-nearest neighbours etc.




Building recommender systems is challenging, because
data is large yet often sparse.
For instance, the ratio of available ratings
vs. all possible user-item valuations for the Netflix Prize
(obviously, it is just a sample of
the complete dataset that Netflix has) is equal to:

```{r netflix_count}
100480507/(480189*17770)
```

A *sparse matrix* (see R package `Matrix`) data structure is
often used for storing of and computing over such data  effectively.




Note that some users are *biased* in the sense that they are more critical or
enthusiastic than
average users.

{ BEGIN exercise }
Is 3 stars a "bad", "fair enough" or "good" rating for you?
Would you go to a bar/restaurant ranked 3.0 by you favourite Maps app community?
{ END exercise }

It is particularly challenging to predict the preferences of users
that cast few ratings, e.g., those who just signed up (*the cold start problem*).







> "Hill  et  al.  [1995]  have  shown  that  users  provide  inconsistent  ratings  when  asked  to  rate  the  same  movie  at  different  times.  They suggest that an algorithm cannot be more accurate than the variance in a user’s ratings for the same item." [@herlocker_etal: p. 6]

It is good to take into account the temporal (time-based) characteristics of data
as well as external knowledge
(e.g., how long ago a rating was cast,
what is a film's genre).

The presented approaches are vulnerable to attacks -- bots may be used
to promote or inhibit selected items.





### Further Reading

```{r recommenderlab,eval=FALSE,echo=FALSE}
library("recommenderlab")
```

Recommended further reading:
[@herlocker_etal],
[@ricci_etal],
[@lu_etal],
[@movielens].
See also the Netflix prize winners: [@bellkor_netflix],
[@bigchaos_netflix],
[@pragmatictheory_netflix].
Also don't forget to take a look at
the R package `recommenderlab` (amongst others).
