<!--
kate: indent-width 4; word-wrap-column 74; default-dictionary en_AU
Copyright (C) 2020-2021, Marek Gagolewski <https://www.gagolewski.com>
This material is licensed under the Creative Commons BY-NC-ND 4.0 License.
-->

# Matrix Algebra in R


Vectors are one-dimensional objects -- they represent "flat" sequences of values.
Matrices, on the other hand, are two-dimensional -- they represent tabular data,
where values aligned into rows and columns. Matrices (and their extensions --
data frames, which we'll cover in the next chapter) are predominant
in data science, where objects are typically represented by means
of feature vectors.

Below are some examples of structured datasets in matrix forms.


```{r intro1}
head(as.matrix(iris[,1:4]))
```

```{r intro2}
WorldPhones
```

The aim of this chapter is to cover the most essential matrix operations,
both from the computational perspective and the mathematical one.






## Creating Matrices


### `matrix()`

A matrix can be created -- amongst others -- with a call to the `matrix()`
function.


```{r mat1}
(A <- matrix(c(1, 2, 3, 4, 5, 6), byrow=TRUE, nrow=2))
class(A)
```

Given a numeric vector of length 6, we've asked R to convert to
a numeric matrix with 2 rows (the `nrow` argument).
The number of columns has been deduced automatically
(otherwise, we would additionally have to  pass `ncol=3` to the function).



Using mathematical notation,
above we have defined $\mathbf{A}\in\mathbb{R}^{2\times 3}$:

\[
\mathbf{A}=
\left[
\begin{array}{ccc}
a_{1,1} & a_{1,2} & a_{1,3} \\
a_{2,1} & a_{2,2} & a_{2,3} \\
\end{array}
\right]
=
\left[
\begin{array}{ccc}
1 & 2 & 3 \\
4 & 5 & 6 \\
\end{array}
\right]
\]



We can fetch the size of the matrix by calling:

```{r mat2}
dim(A) # number of rows, number of columns
```


We can also "promote" a "flat" vector to a column vector, i.e.,
a matrix with one column by calling:

```{r mat3}
as.matrix(1:3)
```





### Stacking Vectors


Other ways to create a matrix involve stacking a couple of vectors
of equal lengths along each other:


```{r mat4}
rbind(1:3, 4:6, 7:9) # row bind
cbind(1:3, 4:6, 7:9) # column bind
```

These functions also allow for adding new rows/columns to existing matrices:

```{r mat6}
rbind(A, c(-1, -2, -3))
cbind(A, c(-1, -2))
```


<!--
Not here: Reading Data

read csv

scan

json

xml
-->



### Beyond Numeric Matrices

Note that logical matrices are possible as well.
For instance, knowing that comparison such as `<` and `==`
are performed elementwise also in the case of matrices, we can obtain:

```{r logimat}
A >= 3
```

Moreover, although much more rarely used, we can define character
matrices:

```{r charmat}
matrix(LETTERS[1:12], ncol=6)
```



### Naming Rows and Columns

Just like vectors could be equipped with `names` attribute:

```{r names1}
c(a=1, b=2, c=3)
```

matrices can be assigned row and column labels
in the form of a list of two character vectors:

```{r names2}
dimnames(A) <- list(
    c("a", "b"),     # row labels
    c("x", "y", "z") # column labels
)
A
```

### Other Methods

The `read.table()` (and its special case, `read.csv()`),
can be used to read a matrix from a text file.
We will cover it in the next chapter, because technically
it returns a data frame object (which we can convert to a matrix with a call
to `as.matrix()`).

`outer()` applies a given (vectorised) function
on each pair of elements from two vectors, forming a two-dimensional "grid".
More precisely `outer(x, y, f, ...)` returns a matrix $\mathbf{Z}$ with
`length(x)`
rows and `length(y)` columns such that $z_{i,j}=f(x_i, y_j, ...)$,
where `...` are optional further arguments to `f`.

```{r outer}
outer(c(1, 10, 100), 1:5, "*") # apply the multiplication operator
outer(c("A", "B"), 1:8, paste, sep="-") # concatenate strings
```


`simplify2array()` is an extension of the `unlist()` function.
Given a list of vectors, each of length one, it will return an "unlisted" vector.
However, if a list of equisized vectors of greater lengths is given,
these will be converted to a matrix.

```{r simplify}
simplify2array(list(1, 11, 21))
simplify2array(list(1:3, 11:13, 21:23))
simplify2array(list(1, 11:12, 21:23)) # no can do
```

`sapply(...)` is a nice application of the above, meaning `simplify2array(lapply(...))`.

```{r sapply}
sapply(split(iris$Sepal.Length, iris$Species), mean)
sapply(split(iris$Sepal.Length, iris$Species), summary)
```

Of course, custom functions can also be applied:

```{r min_mean_max}
min_mean_max <- function(x) {
    # returns a named vector with three elements
    # (note that the last expression in a function's body
    #  is its return value)
    c(min=min(x), mean=mean(x), max=max(x))
}
sapply(split(iris$Sepal.Length, iris$Species), min_mean_max)
```


Lastly, `table(x, y)` creates a contingency matrix that
counts the number of unique pairs of corresponding elements
from two vectors of equal lengths.

```{r table}
library("titanic") # data on the passengers of the RMS Titanic
table(titanic_train$Survived)
table(titanic_train$Sex)
table(titanic_train$Survived, titanic_train$Sex)
```


### Internal Representation (\*)


Note  that by setting `byrow=TRUE` in a call to the `matrix()` function above,
we are reading the elements
of the input vector in the row-wise (row-major) fashion.
The default is the column-major order, which might be a little unintuitive
for some of us.

```{r internal1}
A <- matrix(c(1, 2, 3, 4, 5, 6), ncol=3, byrow=TRUE)
B <- matrix(c(1, 2, 3, 4, 5, 6), ncol=3) # byrow=FALSE
```


It turns out that is exactly the order in which the matrix is stored internally.
Under the hood, it is an ordinary numeric vector:

```{r internal2}
mode(B)    # == mode(A)
length(B)  # == length(A)
as.numeric(A)
as.numeric(B)
```

Also note that we can create a different *view* on the same underlying
data vector:

```{r internal3}
dim(A) <- c(3, 2) # 3 rows, 2 columns
A
dim(B) <- c(3, 2) # 3 rows, 2 columns
B
```







## Common Operations

### Matrix Transpose

The matrix *transpose* is denoted with $\mathbf{A}^T$:

```{r trans}
t(A)
```

Hence, $\mathbf{B}=\mathbf{A}^T$ is a matrix such that $b_{i,j}=a_{j,i}$.

In other words, in the transposed matrix, rows become columns and columns become rows.
For example:

\[
\mathbf{A}=
\left[
\begin{array}{ccc}
a_{1,1} & a_{1,2} & a_{1,3} \\
a_{2,1} & a_{2,2} & a_{2,3} \\
\end{array}
\right]
\qquad
\mathbf{A}^T=
\left[
\begin{array}{cc}
a_{1,1} & a_{2,1} \\
a_{1,2} & a_{2,2} \\
a_{1,3} & a_{2,3} \\
\end{array}
\right]
\]








### Matrix-Scalar Operations


Operations such as $s\mathbf{A}$ (multiplication of a matrix
by a scalar), $-\mathbf{A}$, $s+\mathbf{A}$ etc.
are applied on each element of the input matrix:

```{r matscal1}
(A <- matrix(c(1, 2, 3, 4, 5, 6), byrow=TRUE, nrow=2))
(-1)*A
```


In R, the same rule holds when we compute other
operations (despite the fact that, mathematically,
e.g., $\mathbf{A}^2$ or $\mathbf{A}\ge 0$ might have a different meaning):

```{r matscal2}
A^2 # this is not A-matrix-multiply-A, see below
A>=3
```


### Matrix-Matrix Operations


If $\mathbf{A},\mathbf{B}\in\mathbb{R}^{n\times p}$
are two matrices of identical sizes, then
$\mathbf{A}+\mathbf{B}$ and
$\mathbf{A}-\mathbf{B}$ are understood elementwise,
i.e., they result in $\mathbf{C}\in\mathbb{R}^{n\times p}$
such that $c_{i,j}=a_{i,j}\pm b_{i,j}$.

```{r matmat1}
A-A
```



In R (but not when we use mathematical notation),
all other arithmetic, logical and comparison operators are also
applied in an elementwise fashion.

```{r matmat2}
A*A
(A>2) & (A<=5)
```


### Matrix Multiplication (\*)



Mathematically, $\mathbf{A}\mathbf{B}$
denotes the **matrix multiplication**. It is a very different operation
to the elementwise multiplication.


```{r matmat3}
(A <- rbind(c(1, 2), c(3, 4)))
(I <- rbind(c(1, 0), c(0, 1)))
A %*% I # matrix multiplication
```

This is not the same as the elementwise `A*I`.



Matrix multiplication can only be performed on two matrices of
*compatible sizes* -- the number of columns in the left matrix must match
the number of rows in the right operand.

Given $\mathbf{A}\in\mathbb{R}^{n\times p}$
and $\mathbf{B}\in\mathbb{R}^{p\times m}$, their multiply is a matrix
$\mathbf{C}=\mathbf{A}\mathbf{B}\in\mathbb{R}^{n\times m}$
such that $c_{i,j}$ is the dot product of the $i$-th row in $\mathbf{A}$
and the $j$-th column in $\mathbf{B}$:
\[
c_{i,j} = \mathbf{a}_{i,\cdot} \cdot \mathbf{b}_{\cdot,j}
= \sum_{k=1}^p a_{i,k} b_{k, j}
\]
for $i=1,\dots,n$ and $j=1,\dots,m$.


{ BEGIN exercise }
Multiply
a few simple matrices of sizes $2\times 2$, $2\times 3$, $3\times 2$ etc.
using pen and paper and checking the results in R.
{ END exercise }

<!-- TODO: graphical interpretation -->



Also remember that, mathematically,
*squaring* a matrix is done in terms of matrix multiplication,
i.e., $\mathbf{A}^2 = \mathbf{A}\mathbf{A}$.
It can only be performed on *square* matrices, i.e., ones with the same number
of rows and columns.
This is again different than R's elementwise `A^2`.



Note that $\mathbf{A}^T \mathbf{A}$
gives the matrix that consists of the dot products of all the pairs
of columns in $\mathbf{A}$.

```{r crossprod}
crossprod(A) # same as t(A) %*% A
```

In one of the chapters on Regression, we note
that the Pearson linear correlation coefficient
can be beautifully expressed this way.




<!-- TODO? Matrix inverse?? Do we need it? we might for linear regression, but it's already too much -->



### Aggregation of Rows and Columns

The `apply()` function may be used to transform or summarise
individual rows or columns in a matrix. More precisely:

* `apply(A, 1, f)` applies a given function $f$ on each *row* of $\mathbf{A}$.
* `apply(A, 2, f)` applies a given function $f$ on each *column* of $\mathbf{A}$.

Usually, either $f$ returns a single value (when we wish to aggregate
all the elements in a row/column) or returns the same number of values
(when we wish to transform a row/column). The latter case is covered in the next
subsection.

Let's compute the mean of each row and column in A:

```{r apply1}
(A <- matrix(1:18, byrow=TRUE, nrow=3))
apply(A, 1, mean) # synonym: rowMeans(A)
apply(A, 2, mean) # synonym: colMeans(A)
```

We can also fetch the minimal and maximal value by means of the `range()` function:

```{r apply2}
apply(A, 1, range)
apply(A, 2, range)
```

Of course, a custom function can be provided as well.
Here we compute the minimum, average and maximum of each row:

```{r apply3}
apply(A, 1, function(row) c(min(row), mean(row), max(row)))
```

### Vectorised Special Functions

The special functions mentioned in the previous chapter, e.g.,
`sqrt()`, `abs()`, `round()`, `log()`, `exp()`, `cos()`, `sin()`,
are also performed in an elementwise manner when applied on a matrix object.

```{r special1}
round(1/A, 2) # rounds every element in 1/A
```


An example plot of the absolute values of sine and cosine functions
depicted using the `matplot()` function (see Figure \@ref(fig:matplot-sincos)).

```{r matplot-sincos,fig.cap="Example plot with `matplot()`"}
x <- seq(-2*pi, 6*pi, by=pi/100)
Y <- cbind(sin(x), cos(x)) # a matrix with two columns
Y <- abs(Y) # take the absolute value of every element in Y
matplot(x, Y, type="l")
```



### Matrix-Vector Operations

Mathematically, there is no generally agreed upon
convention defining arithmetic operations between matrices and vectors.

> (\*) The only exception is the  matrix -- vector multiplication in the case
where an argument is a column or a row vector, i.e., in fact, a matrix.
Hence, given $\mathbf{A}\in\mathbb{R}^{n\times p}$ we may write
$\mathbf{A}\mathbf{x}$
only if $\mathbf{x}\in\mathbb{R}^{p\times 1}$ is a column vector.
Similarly, $\mathbf{y}\mathbf{A}$ makes only sense
whenever $\mathbf{y}\in\mathbb{R}^{1\times n}$ is a row vector.


Remark.

: Please take notice of the fact that we consistently
discriminate between different bold math fonts and letter cases:
$\mathbf{X}$ is a matrix, $\mathbf{x}$ is a row or column vector
(still a matrix, but a sequence-like one)
and $\boldsymbol{x}$ is an ordinary vector (one-dimensional sequence).


<!--
> This is why, e.g., the $i$-th row of $\mathbf{X}$ is denoted
with $\mathbf{x}_{i,\cdot}$ -- to emphasise that it is a row vector.
-->


<!-- in R an object of type vector gets promoted to a column vector, i.e., a matrix with 1 column -->




However, in R, we might sometimes wish to vectorise
an arithmetic operation between a matrix and a vector in a row- or column-wise
fashion.
For example, if $\mathbf{A}\in\mathbb{R}^{n\times p}$ is a matrix
and $\mathbf{m}\in\mathbb{R}^{1\times p}$ is a row vector,
we might want to subtract $m_i$ from each element in the $i$-th column.
Here, the `apply()` function comes in handy again.



Example: to create a *centred* version of a given matrix,
we need to subtract from each element the arithmetic mean of its column.

```{r centre1}
(A <- cbind(c(1, 2), c(2, 4), c(5, 8)))
(m <- apply(A, 2, mean)) # same as colMeans(A)
t(apply(A, 1, function(r) r-m)) # note the transpose here
```


The above is equivalent to:

```{r centre2}
apply(A, 2, function(c) c-mean(c))
```




## Matrix Subsetting


Example matrices:

```{r select1}
(A <- matrix(1:12, byrow=TRUE, nrow=3))
B <- A
dimnames(B) <- list(
    c("a", "b", "c"),     # row labels
    c("x", "y", "z", "w") # column labels
)
B
```


### Selecting Individual Elements

Matrices are two-dimensional structures: items are aligned in rows and columns.
Hence, to extract an element from a matrix, we will need two indices.
Mathematically, given a matrix $\mathbf{A}$, $a_{i,j}$ stands for the element
in the $i$-th row and the $j$-th column. The same in R:

```{r select2}
A[1, 2] # 1st row, 2nd columns
B["a", "y"] # using dimnames == B[1,2]
```

### Selecting Rows and Columns

We will sometimes use the following notation to emphasise that
a matrix $\mathbf{A}$ consists of $n$ rows
or $p$ columns:

\[
\mathbf{A}=\left[
\begin{array}{c}
\mathbf{a}_{1,\cdot} \\
\mathbf{a}_{2,\cdot} \\
\vdots\\
\mathbf{a}_{n,\cdot} \\
\end{array}
\right]
=
\left[
\begin{array}{cccc}
\mathbf{a}_{\cdot,1} &
\mathbf{a}_{\cdot,2} &
\cdots &
\mathbf{a}_{\cdot,p} \\
\end{array}
\right].
\]


Here, $\mathbf{a}_{i,\cdot}$ is a *row vector* of length $p$,
i.e., a $(1\times p)$-matrix:

\[
\mathbf{a}_{i,\cdot} = \left[
\begin{array}{cccc}
a_{i,1} &
a_{i,2} &
\cdots &
a_{i,p} \\
\end{array}
\right].
\]

Moreover, $\mathbf{a}_{\cdot,j}$ is a *column vector* of length $n$,
i.e., an $(n\times 1)$-matrix:

\[
\mathbf{a}_{\cdot,j} = \left[
\begin{array}{cccc}
a_{1,j} &
a_{2,j} &
\cdots &
a_{n,j} \\
\end{array}
\right]^T=\left[
\begin{array}{c}
{a}_{1,j} \\
{a}_{2,j} \\
\vdots\\
{a}_{n,j} \\
\end{array}
\right],
\]


We can extract individual rows and columns from a matrix by using
the following notation:

```{r select3}
A[1,] # 1st row
A[,2] # 2nd column
B["a",] # of course, B[1,] is still legal
B[,"y"]
```

Note that by extracting a single row/column, we get an atomic (one-dimensional)
vector. However, we can preserve the dimensionality of the output object
by passing `drop=FALSE`:

```{r select4}
A[  1,    , drop=FALSE] # 1st row
A[   ,   2, drop=FALSE] # 2nd column
B["a",    , drop=FALSE]
B[   , "y", drop=FALSE]
```

Now this is what we call proper row and column vectors!



### Selecting Submatrices

To create a sub-block of a given matrix we pass two indexers,
possibly of length greater than one:

```{r select5}
A[1:2, c(1, 2, 4)] # rows 1,2 columns 1,2,4
B[c("a", "b"), c(1, 2, 4)]
A[c(1, 3), 3]
A[c(1, 3), 3, drop=FALSE]
```

### Selecting Based on Logical Vectors and Matrices


We can also subset a matrix with a logical matrix of the same size.
This always yields a (flat) vector in return.

```{r select6}
A[A>8]
```

Logical vectors can also be used as indexers:

```{r select7}
A[c(TRUE, FALSE, TRUE),] # select 1st and 3rd row
A[,colMeans(A)>6] # columns with means > 6
B[B[,"x"]>1 & B[,"x"]<=9,] # All rows where x is in (1, 9]
```


### Selecting Based on Two-Column Matrices

Lastly, note that we can also index a matrix `A`
with a 2-column matrix `I`, i.e., `A[I]`.
This allows for an easy access to
`A[I[1,1], I[1,2]]`, `A[I[2,1], I[2,2]]`, `A[I[3,1], I[3,2]]`, ...

```{r select8}
I <- cbind(c(1, 3, 2, 1, 2),
           c(2, 3, 2, 1, 4)
)
A[I]
```

This is exactly
` `r paste(sprintf("A[%d, %d]", I[,1], I[,2]), collapse=", ")` `.




<!--

TODO: Add OnTrack exercises

-->





## Further Reading

Recommended further reading: [@Rintro]

Other: [@mml], [@rprogdatascience], [@r4ds]
